Oncogenic and teratogenic effects of p53Y217C, a mouse model of the human hotspot mutant p53Y220C

Version 1

Abstract

Abstract

Missense “hotspot” mutations localized in six p53 codons account for 20% of TP53 mutations in human cancers. Hotspot p53 mutants have lost the tumor suppressive functions of the wildtype protein, but whether they may gain additional functions promoting tumorigenesis remains controversial. Here we generated Trp53Y217C, a mouse model of the human hotspot mutant TP53Y220C. DNA damage responses were lost in p53Y217C/Y217C cells. Surprisingly, intercrosses from p53+/Y217C heterozygotes yielded only one p53Y217C/Y217C female for nineteen p53Y217C/Y217C males at weaning, a skewed distribution explained by the high frequency of p53Y217C/Y217C female embryos with exencephaly and the death of most p53Y217C/Y217C female neonates. Furthermore, parturition was impaired in pregnant p53Y217C/Y217C females. Finally, p53Y217C/Y217C males died earlier than p53-/- males, with more aggressive thymic lymphomas. Together, these data indicate that the p53Y217C mutation not only abrogates wildtype p53 functions, but also exerts additional effects promoting oncogenesis in males and teratogenesis or dystocia in females.

Introduction

Somatic alterations in the TP53 gene, encoding the tumor suppressor p53, are the most common events in human tumors (Hollstein et al., 1991). The p53 protein is a stress sensor stabilized and activated in response to potentially oncogenic insults. In its tetrameric active form, wildtype (WT) p53 can trigger a transcriptional program to induce various responses including cell cycle arrest, senescence, apoptosis or metabolic changes (Beckerman and Prives, 2010). In human cancers, about 75% of all TP53 alterations are missense mutations, and the eight most frequent (“hotspot”) missense mutations (R175H, Y220C, G245S, R248Q, R248W, R273H, R273C and R282W) affect six residues localized in the p53 DNA-binding domain (DBD) and account for 20% of all TP53 mutations (Hainaut and Pfeifer, 2016). The fact that most TP53 mutations are missense substitutions suggested that cells expressing mutant p53 might have a selective advantage over cells lacking p53, and evidence for this was first gained by expressing various p53 mutants in p53-null cells, and observing enhanced tumorigenic potential in nude mice or increased plating efficiency in agar cell culture (Dittmer et al., 1993). Mouse models expressing hotspot p53 mutants next helped to define potential mechanisms accounting for accelerated tumorigenesis. First, a dominant-negative effect (DNE) may be observed in heterozygotes, if the mutant p53 inhibits the WT p53 protein in hetero-tetramers (Gencel- Augusto and Lozano, 2020). Evidence that this mechanism accounts for accelerated tumorigenesis was notably reported in human leukemia (Boettcher et al., 2019). A second mechanism is a gain-of-function (GOF), i.e. the acquisition by mutant p53 of new oncogenic properties (Amelio and Melino, 2020; Stein et al., 2020). Although the DNA sequence specificity of p53 mutants is impaired, leading to a loss of WT functions (LOF), many p53 mutant proteins are stabilized in the cell and might engage in aberrant interactions with other transcription factors or chromatin-modifying complexes, leading to the acquisition of GOF phenotypes (Kim and Deppert, 2007; Pfister and Prives, 2017). However, a few recent studies challenged the pathological importance of mutant p53 GOF, or at least its relevance in anti-cancer therapeutic strategies (Aubrey et al., 2018; Boettcher et al., 2019; Wang et al., 2023). An alternative hypothesis emerged, postulating that the tumorigenic properties of a p53 mutant might result from a separation-of-function (SOF), if the mutant retains pro-proliferative or pro-survival functions of WT p53 while losing its tumor suppressive activities (Kennedy and Lowe, 2022). The concept of SOF was first proposed for TP53 exon-6 truncating mutations, which mimic a naturally occurring alternative p53 splice variant (Shirole et al., 2016), but may apply to some p53 missense mutations, for example p53R248W (Humpton et al., 2018). Importantly, the analysis of various mutant p53 alleles in vivo appears crucial for a better understanding of their contribution to tumorigenesis, because GOF/SOF phenotypes may vary depending on the mutated residue, its specific mutation, the cellular context or genetic background (Dibra et al., 2024; Hanel et al., 2013; Kadosh et al., 2020; Kim and Lozano, 2018; Kotler et al., 2018; McCann et al., 2022; Xiong et al., 2022).

In human cancers, the TP53Y220C mutation is the most frequent missense mutation outside of the DNA-binding surface of p53 (Hainaut and Pfeifer, 2016). The somatic TP53Y220C mutation accounts for over 100,000 new cancer cases per year worldwide, and a germline TP53Y220Cmutation was reported in 15 families with the Li-Fraumeni syndrome (LFS) of cancer predisposition (Bouaoun et al., 2016). This mutation causes a structural change in the DBD localized at the periphery of the β−sandwich region of the protein, far from the surface contact of DNA. The mutation from a tyrosine to a cysteine markedly lowers the thermodynamic stability of the DBD, leading to a largely unfolded and inactive protein at body temperature (Joerger et al., 2006), and a molecule designed to bind p53Y220C and restore WT protein conformation appears as a promising anti-cancer drug (Dumbrava et al., 2022). Analyses of the impact of p53Y220C in cancer cell lines led to conflicting results. On one hand, overexpression of p53Y220C in a p53-null cell line increased its capacity for migration or invasion (Zhou et al., 2022), and a decreased expression of p53Y220C, caused by RNA interference or various chemical compounds, promoted cell death or decreased the migratory or invasive capacities of cells (Tseng et al., 2022; Vikhanskaya et al., 2007; Zhou et al., 2022). On the other hand, the removal by CRISPR/Cas9 of p53 mutants with reported GOF - including p53Y220C - in diverse cancer cell lines had no significant impact on cell survival or proliferation, and the cell death caused by RNA interference against p53Y220C might result from non-specific toxic effects (Wang et al., 2023). Here, to gain a better understanding of the impact of the p53Y220C mutation in vivo, we generated a mouse model with a targeted Trp53Y217C mutation - the mouse equivalent to human TP53Y220C (Figure S1) - and analyzed animals and cells carrying one or two copies of the mutant allele.

Results

Targeting of a p53Y217C mutation in the mouse

We used homologous recombination in 129/SvJ ES cells to target the p53Y217C mutation at the mouse Trp53 locus. Mice expressing p53Y217C conditionally were generated by using a targeting vector containing transcriptional stops flanked by LoxP sites (LSL) upstream of coding sequences, and the p53Y217C mutation in exon 6 (Figure 1A-D). p53+/LSL-Y217C mice were then crossed with mice carrying the PGK-Cre transgene (Lallemand et al., 1998) to excise the LSL cassette and obtain p53+/Y217C mice, expressing the mutant protein constitutively. After two backcrosses over a C57BL/6J background, mouse embryonic fibroblasts (MEFs) were prepared from intercrosses of p53+/Y217C mice (Figure 1E). We extracted RNAs from p53Y217C/Y217C MEFs then sequenced p53 mRNAs to verify that the p53Y217C coding sequence was identical to the wildtype p53 (p53WT) sequence, except for the desired missense mutation and a silent mutation introduced to facilitate mouse genotyping (Figure 1F). The quantification of p53 mRNA levels from WT and p53Y217C/Y217C MEFs next revealed similar transcription from both alleles (Figure 1G).

Targeting the Y217C missense mutation at the mouse Trp53 locus.

(A) Targeting strategy. The Trp53 gene is within a 17 kb-long EcoRI (RI) fragment (black boxes are for coding sequences and white boxes for UTRs). The targeting construct contains: (1) a 1.5 kb-long 5’ homology region; (2) a Lox-Stop-Lox (LSL) cassette with a neomycin selection gene (Neo), four transcriptional stops (STOP) and an EcoRI site, flanked by LoxP sites (arrowheads); (3) p53 exons, including the Y217C (YC) missense mutation in exon 6 (asterisk) and an additional BanII site; (4) a 2.8 kb-long 3’ homology region; and (5) the diphteria α-toxin (DTA) gene for targeting enrichment. Proper recombinants with a p53LSL-Y217C allele were identified by a 2.4 kb-long band after PCR with primers a and b, and confirmed by bands of 635 and 224 bp after PCR with primers c and d and BanII digestion. They were also verified by Southern blot with the indicated probe as containing a 10.5 kb EcoRI band. Two recombinant ES clones were injected into blastocysts to generate chimeras, and germline transmission was verified by genotyping with primers c and d and BanII digestion. Excision of the LSL cassette was performed in vivo, by breeding p53+/LSL-Y217C male mice with females carrying the PGK-Cre transgene. (B-D) Screening of recombinant ES clones (+) by PCR with primers a and b (B); PCR with primers c and d then BanII digestion (C); Southern blot (D). (E) Genotyping of mouse embryonic fibroblasts (MEFs) from an intercross of p53+/Y217C mice, by PCR with primers c and d and BanII digestion. (F) p53Y217C sequence around codon 217. The introduced Y217C missense mutation and the mutation creating an additional BanII restriction site are highlighted (asterisks). (G) WT and p53Y217C/Y217C MEFs express similar p53 mRNA levels. Total RNA was extracted, then p53 mRNAs were quantified by real-time qPCR, normalized to control mRNAs and the amount in WT cells was assigned the value of 1. Primer sequences are listed in Table S6.

Deficient p53-dependent stress responses in p53Y217C/Y217C cells

We used western blots to analyze protein extracts from p53+/+, p53+/-, p53+/Y217C, p53Y217C/Y217C and p53-/- MEFs, unstressed or treated with Nutlin, an antagonist of Mdm2, the E3 ubiquitin ligase for p53 (Vassilev et al., 2004). Results indicated a high increase in p53 abundance in untreated and Nutlin-treated p53Y217C/Y217C MEFs and a moderate increase in untreated and Nutlin-treated p53+/Y217C MEFs, compared to WT cells (Figures 2A and S2). Protein levels for p21 and Mdm2, the products of two classical p53 target genes, appeared similar in p53+/+ and p53+/Y217C MEFs, and were undetectable or markedly decreased in p53Y217C/Y217C and p53-/- MEFs (Figures 2A and S2). Accordingly, p53Y217C appeared unable to transactivate the Cdkn1a/p21 and Mdm2 genes or to bind their promoters (Figure 2B). The fractionation of cells before or after treatment with doxorubicin, a clastogenic drug, indicated that p53WT accumulated in the nucleoplasm and chromatin fractions in response to DNA damage, whereas p53Y217C appeared mostly cytoplasmic in both untreated and doxorubicin-treated cells, and undetectable or barely detectable in chromatin fractions (Figures 2C and S3). Furthermore, when Nutlin-treated MEFs were analyzed by immunofluorescence, p53WT was only detected in nuclei, whereas p53Y217C could be observed in nuclear and cytoplasmic compartments (Figure 2D). We next analyzed two well-known p53-mediated responses to DNA damage, i.e. cell cycle arrest in MEFs and apoptosis in thymocytes (Kastan et al., 1992; Lowe et al., 1993). WT MEFs exposed to increasing doses of ɣ-irradiation (3 Gy or 12 Gy) exhibited significant increases in G1/S ratios, whereas G1/S ratios were similar before or after irradiation in p53Y217C/Y217C and p53-/- MEFs (Figures 2E and S4). Furthermore, thymocytes recovered from irradiated WT mice underwent a massive apoptotic response, with an almost three-fold increase in apoptotic cells after a 10 Gy irradiation. By contrast, no increase in apoptotic cells was observed upon irradiation in the thymi from p53Y217C/Y217C mice, and apoptotic thymocytes were equally rare in irradiated p53Y217C/Y217C and p53-/- mice (Figures 2F and S5). Together, these results indicated that the p53Y217C mutation altered the abundance and intracellular distribution of the p53 protein, associated with a decrease in DNA-bound protein, and that p53Y217C had completely lost the ability to induce cell cycle arrest and apoptosis upon DNA damage.

p53Y217C/Y217C cells exhibit alterations in p53 abundance and subcellular localization and defective responses to DNA damage.

(A) Increased p53 protein levels in p53Y217C/Y217C and p53+/Y217C MEFs. MEFs of the indicated genotypes were treated or not with 10 μM Nutlin 3a for 24 h, then protein extracts were immunoblotted with antibodies against Mdm2, p53, p21 and actin. (B) The transactivation of classical p53 target genes Cdkn1a/p21 and Mdm2 is impaired in p53Y217C/Y217C cells. WT, p53Y217C/Y217C and p53-/- MEFs were treated as in (A), then (top) mRNAs were quantified in 5-6 independent experiments using real-time PCR, with results normalized to control mRNAs and mean RNA amounts in unstressed p53+/+ cells assigned a value of 1; or (bottom) ChIP assays were performed at the Cdkn1a and Mdm2 promoters in 2-3 independent experiments with an antibody against p53 or rabbit IgG as a negative control. Immunoprecipitates were quantified using real-time PCR, normalized to data over an irrelevant region, and the amount in unstressed p53+/+ cells was assigned a value of 1. (C) Assessment of p53WT and p53Y217C subcellular localization by cellular fractionation. WT and p53Y217C/Y217C MEFs were treated or not with doxorubicin for 24 hr, submitted to cellular fractionation, then protein extracts were immunoblotted with antibodies against p53 or the fraction controls Tubulin for cytoplasm (Cp.), Nup98 for nucleoplasm (Np.) and histone H3 for chromatin (χin). (D) Assessment of p53WT and p53Y217C subcellular localization by immunofluorescence. WT, p53Y217C/Y217C and p53-/- MEFs were treated with 10 μM Nutlin 3a for 24 h, then stained with antibodies against p53 (red) or actin (green) and DNA was counterstained with DAPI (blue). (E) Absence of a cell cycle arrest response in p53Y217C/Y217C MEFs. Asynchronous cell populations of p53+/+, p53Y217C/Y217C and p53-/- MEFs were analyzed 24 hr after 0, 3 or 12 Gy ɣ-irradiation. Results from 3 independent experiments. (F) Absence of a p53-dependent apoptotic response in p53Y217C/Y217C thymocytes. Age-matched mice of the indicated genotypes were left untreated or submitted to 10 Gy whole-body ɣ-irradiation then sacrificed after 4 hr and their thymocytes were stained with Annexin V-FITC and analyzed by FACS, in 2 independent experiments. ***P<0.001, **P<0.01, *P<0.05, °P=0.09, ns: non-significant by Student’s t tests.

Impact of p53Y217C on mouse development and parturition

We next determined the impact of the p53Y217C mutation in vivo, by comparing mouse cohorts generated from intercrosses of heterozygous p53+/- or p53+/Y217C mice resulting from ≥ 5 backcrosses to the C57BL6/J background. Intercrosses of p53+/- mice are known to yield a reduced proportion of p53-/- mice, that is mainly due to defects in neural tube closure causing exencephaly in a fraction of p53-/- female embryos (Armstrong et al., 1995; Sah et al., 1995) and, to a lesser extent, to lung dysfunction affecting a fraction of p53-/- neonates (Tateossian et al., 2015). At weaning (on the 21st day postpartum or P21), we observed one p53-/- female mouse for 3.5 p53-/- males from p53+/- intercrosses, an underrepresentation of females consistent with frequencies reported in earlier studies (Armstrong et al., 1995; Sah et al., 1995). Strikingly, the underrepresentation of female mice at weaning age was even more acute for p53Y217C/Y217C animals, with only one p53Y217C/Y217C female for 19 p53Y217C/Y217C males from p53+/Y217C intercrosses (Figure 3A). We next analyzed p53Y217C/Y217C embryos, generated from heterozygous intercrosses, or from heterozygous-homozygous (p53+/Y217C x p53Y217C/Y217C) crosses, at embryonic days E12.5-E16.5. An underrepresentation of p53Y217C/Y217C female embryos was not observed, but 11/26 (42%) female embryos exhibited developmental abnormalities, including 10 with exencephaly, whereas all the male embryos appeared normal (Figure 3B). Importantly, the frequency of p53Y217C/Y217C female embryos with exencephaly (38.5%) was much higher than the reported frequency (0-8%) of p53-/- female embryos with exencephaly of C57BL/6 x 129/Sv genetic background (Donehower et al., 1992; Sah et al., 1995), suggesting a stronger effect of the p53Y217C mutant on female embryonic development.

p53Y217C/Y217C mice exhibit female-specific perinatal lethality.

(A) Distribution of weaned mice obtained from p53+/- or p53+/Y217C intercrosses. Obs: observed numbers of mice at weaning (P21); exp: expected numbers assuming a Mendelian distribution without sex distortion; f/m: observed female/male ratios. Consistent with previous reports the observed distribution of weaned mice from p53+/- intercrosses did not conform to values expected for a Mendelian distribution without sex distortion (υ=5; χ2= 16.3>15.09), indicating a significant deficit in female p53-/- mice (top). The distribution of weaned mice from p53+/Y217C intercrosses diverged even more from values for a Mendelian distribution without sex distortion (υ =5; χ2= 91.02>15.09), due to a striking deficit in female p53Y217C/Y217C mice (bottom). The female frequency is significantly lower for p53Y217C/Y217C (3/535) than for p53-/- (8/196) females (P=0.0018 in a Fischer’s test). (B) Exencephaly is frequently observed in p53Y217C/Y217C female embryos. Top: the distribution of E12.5-16.5 embryos from heterozygous (p53+/Y217C) intercrosses or heterozygous-homozygous (p53+/Y217C x p53Y217C/Y217C) crosses is shown. f or m exenc.: number of female or male embryos with exencephaly; o.a.: embryos with other abnormalities. Below, examples of female p53Y217C/Y217C embryos at E12.5, 13.5 and 16.5 exhibiting exencephaly (arrows) are each shown (center) together with a normal embryo from the same litter (bottom). (C) Distribution of mice at birth from the indicated crosses.

In p53-/- female mice, exencephaly was previously correlated with stochastic aberrant X chromosome inactivation, with a decreased expression of Xist and an increase in bi-allelic expression of X-linked genes including Maob, Pls3 and Usp9x (Delbridge et al., 2019). We prepared neurospheres from p53+/+, p53-/- and p53Y217C/Y217C female embryos and quantified mRNAs for these genes in neurospheres. Compared to neurospheres from WT animals, p53Y217C/Y217C and p53-/- neurospheres exhibited an apparent decrease in Xist expression and significantly higher levels of Maob, Pls3 and Usp9x (Figure S6). Thus, as for female p53-/- mice, aberrant chromosome X inactivation may contribute to the underrepresentation of p53Y217C/Y217C female mice. Of note however, defects in chromosome X inactivation did not appear to be more pronounced in p53Y217C/Y217C neurospheres compared to p53-/- neurospheres.

To further analyze the impact of the p53Y217C mutation during development, we also analyzed its potential effect in embryos lacking Mdm2 or Mdm4, two major p53 regulators. The embryonic lethality resulting from Mdm2 or Mdm4 loss is rescued by a concomitant p53 deficiency (Bardot et al., 2015; Finch et al., 2002; Jones et al., 1995; Migliorini et al., 2002; Montes de Oca Luna et al., 1995; Parant et al., 2001), which provides a powerful assay for analyzing the functionality of p53 mutant alleles (Bardot et al., 2015; Iwakuma et al., 2004; Marine et al., 2006; Toledo et al., 2006). We identified Mdm2-/- p53Y217C/Y217C and Mdm4-/- p53Y217C/Y217C viable mice of both sexes, consistent with a major loss of canonical WT p53 activities in the p53Y217C mutant (Table S1). Of note, a deficit in homozygous p53Y217C/Y217C females was also observed at weaning among animals lacking Mdm2 or Mdm4. Interestingly however, the female to male ratio for p53Y217C/Y217C animals appeared markedly increased in genetic backgrounds with Mdm4 loss (1/6) or haploinsufficiency (4/15), but only marginally increased (if at all) in genetic backgrounds with Mdm2 loss (1/11) or haploinsufficiency (1/11). This suggested that altering the levels of p53 regulators, particularly Mdm4, might improve the survival of p53Y217C/Y217C females (Table S1).

Importantly, the high frequency (42%) of abnormal p53Y217C/Y217C female embryos at E12.5-E16.5 could only partially account for the acute deficit in p53Y217C/Y217C females observed at weaning, suggesting either embryonic abnormalities undetectable macroscopically, or that an important fraction of p53Y217C/Y217C females died later in development or postpartum. We performed additional crosses to analyze female to male ratios at postpartum day 0 or 1 (P0-P1) and observed 20 females for 18 males for p53+/+ animals, but only 5 females for 22 males for p53Y217C/Y217C animals (Figure 3C). Furthermore, only one of these p53Y217C/Y217C females reached weaning age, and the 4 other neonates died before P2. Altogether, these data led to conclude that the p53Y217C mutation caused female-specific developmental abnormalities at a higher penetrance than a null allele, leading to a perinatal (late embryonic or early postpartum) lethality for most homozygous mutant females.

Interestingly, we observed that p53Y217C/Y217C males were fertile, and included them in some of our crosses to generate homozygous mutants at higher frequencies (Figure 3B-C and Table S1). By contrast, we were able to mate two p53Y217C/Y217C adult females with a p53+/Y217C male, and both got pregnant but had to be sacrificed due to complications while giving birth. In both cases, extended labor (> 24 hr) was the main sign of dystocia (Table S2). These observations suggest that, even for the rare p53Y217C/Y217C females able to reach adulthood, the p53Y217C mutation caused pathological traits not observed in p53-/- females, because the loss of p53 was not previously reported to cause dystocia (Embree-Ku and Boekelheide, 2002; Guimond et al., 1996; Hu et al., 2007). Together, our data indicated that the p53Y217C mutation not only abolished canonical p53 activities, but also conferred additional effects compromising female perinatal viability or parturition.

Impact of p53Y217C on tumorigenesis

We next analyzed the impact of the p53Y217C mutation on the onset and spectrum of spontaneous tumors in mice. We first compared p53+/Y217C, p53+/- and p53+/+ mouse cohorts for 2 years. p53+/Y217C and p53+/- cohorts exhibited similar spontaneous tumor onset and spectrum, with most mice developing sarcomas during their second year of life (Figure S7). Thus, at least for tumors arising spontaneously, the p53Y217C mutant protein did not appear to exert a dominant negative effect over the wildtype p53 protein.

We next compared p53Y217C/Y217C and p53-/- mouse cohorts. Because of the difficulty to generate p53Y217C/Y217C females, we restricted our comparison to males, to avoid potential biases that might result from different sex ratios. All p53-/- mice are known to die in less than 1 year, from thymic lymphomas in most cases, or more rarely from sarcomas (Donehower et al., 1992; Jacks et al., 1994). p53Y217C/Y217C males died faster than their p53-/- counterparts: all the p53Y217C/Y217C males were dead by the age of 7 months, whereas more than 20% of the p53-/- males were still alive at that age (Figure 4A). Furthermore, most p53-/- and p53Y217C/Y217C mice died from thymic lymphomas, but histological analysis of tumor organs revealed that the lymphomas in p53Y217C/Y217C males were more aggressive and invasive, with sites of metastases notably including the lungs and spleen (Figure 4B-C). Thus, the p53Y217C mutation not only abolished canonical tumor suppressive activities but also apparently conferred oncogenic properties to the encoded protein. Consistent with this, when we performed an RNA-seq analysis of thymi from 8 weeks-old p53+/+, p53-/- and p53Y217C/Y217C males, 81.5 % of the 717 differentially expressed genes indicated a loss of function (LOF) in the p53Y217C mutant, but genes suggesting a gain of function (GOF) or a separation of function (SOF) were also observed (Figure S8A and Table S3). Of note, among the differentially expressed genes corresponding to a LOF in the p53Y217C mutant were Bbc3/Puma, Cdkn1a/p21 and Zmat3 (Figure S8B), three p53-target genes known to play a major role in p53-mediated tumor suppression (Brennan et al., 2023).

Oncogenic effects of the mutant protein in p53Y217C/Y217C male mice. (A-B)

In homozygous males, p53Y217C leads to accelerated tumor onset and death (A), and aggressive metastatic tumors (B); n=cohort size. (C) Hematoxylin and eosin (H&E) staining of sections from the lung (top) and spleen (bottom) of p53-/- and p53Y217C/Y217C male mice, showing metastases in p53Y217C/Y217C animals. Normal organ structures are shown, with “A” indicating pulmonary alveoli, and “WP” and “RP” standing for splenic white and red pulp, respectively. In the lung section of the p53Y217C/Y217C mouse, the rectangle indicates a lymphoma area. In the spleen section of the p53Y217C/Y217C mouse, the typical splenic structures are absent due to massive tissue homogenization of the spleen by lymphoma cells.

We next performed a comparative analysis focusing on p53Y217C/Y217C and p53-/- thymi, which revealed 192 differentially expressed genes (Figure 5A-B, Table S4). An analysis of these data with GOrilla, the Gene Ontology enRIchment anaLysis and visuaLizAtion tool (Eden et al., 2009), indicated that 141 of these genes were associated with at least one gene ontology (GO) term, and revealed a significant enrichment for genes associated with white blood cell chemotaxis/migration and inflammation (Figure 5C and Figure S9). Among these genes were notably Ccl17, Ccl9, Ccr3, Cxcl10, S100a8 and S100a9, six genes associated each with 10-15 GO terms related to white blood cell behavior or inflammation (Table S5). We next used RT- qPCR to quantify the expression of these genes in thymi from 8 weeks-old p53+/+, p53Y217C/Y217C and p53-/- male mice, and found their expression to be significantly increased in p53Y217C/Y217C thymi compared to p53-/-, or to both p53+/+ and p53-/- mice, consistent with gain or separation of function (GOF/SOF) effects in the p53Y217C mutant (Figure 5D). Furthermore, when we compared the transcriptomes of p53Y217C/Y217C and p53-/- thymocytes by gene set enrichment analysis (Subramanian et al., 2005), we found 13 gene sets significantly enriched in p53Y217C/Y217C cells (with normalized enrichment scores > 2), among which 3 (“antimicrobial peptides”, “chemokine receptors bind chemokines” and “defensins”) were related to an immune response (Figures 5E and S10A), consistent with an inflammatory response associated with the p53Y217C mutant. Other gene sets significantly enriched in p53Y217C/Y217C cells notably included 5 sets (“electron transport chain”, “respiratory electron transport ATP synthesis by chemiosmotic coupling and heat production by uncoupling proteins”, “mitochondrial translation”, “respiratory electron transport” and “oxidative phosphorylation”) related to mitochondria function, and 3 sets (“deposition of new CENPA-containing nucleosomes at the centromere”, “arginine methyltransferases methylate histone arginines” and “PRC2 methylates histones and DNA”) related to chromatin plasticity (Figure 5E and S10B-C). These gene sets, differentially expressed between p53Y217C/Y217C and p53-/- cells, might contribute to accelerated tumorigenesis in p53Y217C/Y217C mice, given the reported impact of inflammation, metabolism changes and epigenetic plasticity on cancer evolution (Feinberg et al., 2006; Greten and Grivennikov, 2019; Kroemer and Pouyssegur, 2008).

Increased expression of genes associated with white blood cell chemotaxis and inflammation in p53Y217C/Y217C male thymi. (A-B)

RNAseq analysis of thymi from p53Y217C/Y217C (n=3) and p53-/- (n=3) 8 weeks-old male mice. Volcano plot (A), with differentially expressed genes (DEGs), downregulated (blue) or upregulated (red) in p53Y217C/Y217C cells. Unsupervised clustering heat-map plot (B), with 192 DEGs ranked according to log2 fold changes. (C) Gene Ontology (GO) analysis of DEGs. Out of 192 DEGs, 141 are associated with at least one GO term, according to the Gene Ontology enRIchment anaLysis and visuaLizAtion tool (GOrilla). For each GO term, enrichment was calculated by comparing with the full list of 19759 genes associated with a GO term, and results are presented here with a color scale according to P-value of enrichment. This analysis mainly revealed an enrichment for genes associated with white blood cell chemotaxis/migration and inflammation, as shown here. Complete results of the analysis are presented in Figure S9. (D) RT-qPCR analysis of the indicated genes in thymi from 8 weeks-old p53+/+, p53Y217C/Y217C and p53-/- male mice. Means + s.e.m. (n=3 per genotype). ***P<0.001, **P<0.01, *P<0.05, °P<0.075 by Student’s t tests. (E) Gene set enrichment analysis (GSEA) of transcriptomes from p53Y217C/Y217C and p53-/- thymic cells. GSEA identified 13 gene sets enriched in p53Y217C/Y217C cells, with normalized enrichment scores (NES) > 2. Nominal p-values (NOM p-val) and false discovery rate q-values (FDR q-val) are indicated.

Discussion

In this report, we generated a mouse model with a Trp53Y217C allele, the murine equivalent of the human hotspot mutant TP53Y220C. The analysis of this mouse model indicated that the p53Y217C mutation leads to the loss of many canonical wildtype p53 activities, notably the capacity to transactivate target genes important for tumor suppression (e.g. Bbc3, Cdkn1a and Zmat3). The fact that p53Y217C can rescue the embryonic lethality caused by a deficiency in Mdm2 or Mdm4 is also consistent with a major loss of canonical wildtype p53 functions in this mutant.

In addition, our analyses provide evidence that the p53Y217C mutation exhibits oncogenic effects. The possibility that a mutant p53 might acquire oncogenic functions was first suggested over 30 years ago (Dittmer et al., 1993), but this notion became controversial in recent years. Indeed, studies of a few p53 mutants in human acute myeloid leukemia and a mouse model of B cell lymphoma indicated that their increased tumorigenicity would result from dominant- negative effects (DNE) rather than gain-of-function (GOF), and a recent study further suggested that the putative GOF of many p53 mutants would not be required to sustain the expansion of tumors (Aubrey et al., 2018; Boettcher et al., 2019; Wang et al., 2023). Here, however, we observed accelerated tumorigenesis in p53Y217C/Y217C mice, but did not observe any evidence of DNE in p53+/Y217C animals. Importantly, the fact that p53Y217C did not exhibit a DNE over the WT protein is consistent with the report that Li-Fraumeni patients carrying an heterozygous p53Y220C mutation display a similar age of cancer onset than Li-Fraumeni patients with an heterozygous non-sense p53 mutation (Xu et al., 2014). Furthermore, our evidence of an oncogenic GOF leading to aggressive metastatic tumors in p53Y217C/Y217C male mice is consistent with the oncogenic GOF recently attributed to the p53Y220C mutant in male patients with glioblastoma (Rockwell et al., 2021), or with experiments suggesting that the expression of p53Y220C in p53-null cells, or its overexpression in MCF-10A cells, may increase their capacity for migration or invasion (Pal et al., 2023; Zhou et al., 2022). Our transcriptomic analyses suggested that the p53Y217C mutant might accelerate tumorigenesis by promoting inflammation, a hallmark of cancer (Hanahan, 2022) previously associated with a few other hotspot p53 mutants with oncogenic GOF (Agupitan et al., 2020; Behring et al., 2019; Ham et al., 2019; Zhao et al., 2024). Interestingly, our GSEA comparison of transcriptomes from p53Y217C/Y217C and p53-/- male thymic cells indicated differences related to inflammation and chemokine signaling, but also to mitochondria function and chromatin plasticity, and a link between inflammation, mitochondrial copper and epigenetic plasticity was recently demonstrated (Solier et al., 2023).

A striking and unexpected effect of the p53Y217C mutation was its severe impact on the perinatal viability of female mice, which led to observe only 1 p53Y217C/Y217C female for 19 p53Y217C/Y217C males at weaning from heterozygous intercrosses. The perinatal lethality of p53Y217C/Y217C females correlated with a high frequency (38.5%) of exencephalic females at E12.5-E16.5 embryonic ages. By comparison, the female to male ratio at weaning was 1/3.5 for p53-/- mice, and only 0-8% of exencephalic female p53-/- embryos were reported in similar genetic backgrounds (Armstrong et al., 1995; Donehower et al., 1992; Sah et al., 1995). These observations provide evidence of teratogenic effects for the p53Y217C mutant protein. Three other mouse p53 models were previously found to cause exencephaly at a higher frequency than a p53 null allele : p53NLS1, with 3 mutations at residues 316-318 affecting a nuclear localization signal and leading to a predominantly cytoplasmic localization of p53NLS1 in most cells (Regeling et al., 2011); p53N236S, a mouse model of p53N239S, a recurrent but uncommon mutant in human cancers (Zhao et al., 2019); and Bim+/- p53-/- mice, combining p53 loss with an haploinsufficiency in the cytoplasmic proapoptotic regulator Bim/Bcl2l11 (Delbridge et al., 2019). Interestingly, the female-specific lethality of p53N236S mice was proposed to result from increased Xist expression, whereas a decrease in Xist expression was observed in Bim+/- p53-/- mice, suggesting distinct causes for neural tube defects (Delbridge et al., 2019; Zhao et al., 2019). We observed decreased Xist expression and an increased expression of X-linked genes in p53Y217C/Y217C females, which is not consistent with the mechanism proposed for exencephaly in the p53N236S mouse model, but might rather reflect a loss of WT p53 function (Delbridge et al., 2019). In addition, the analysis of the p53NLS1 and Bim+/- p53-/- mouse models make it tempting to speculate that, in female p53Y217C/Y217C embryos, the abundance of p53Y217C in the cytoplasm might perturb mitochondria function (Blandino et al., 2020) to further promote neural tube closure defects. Consistent with this possibility, p53 mutant proteins accumulating in the cytoplasm were reported to inhibit autophagy (Morselli et al., 2008), a key determinant for mitochondrial integrity (Rambold and Lippincott-Schwartz, 2011), and autophagy impairment may promote neural tube defects (Fimia et al., 2007; Ye et al., 2020).

A particularly intriguing question is whether common mechanisms might contribute to the teratogenic effects causing the perinatal lethality of p53Y217C/Y217C females and the oncogenic effects accelerating tumorigenesis in p53Y217C/Y217C males. Interestingly, our data indicated that inflammation might promote oncogenesis in p53Y217C/Y217C males, and inflammation was proposed to promote neural tube defects in a few mouse models (Lian et al., 2004; Zhao et al., 2013). Furthermore CD44, which drives inflammation and cancer progression (Solier et al., 2023), was recently identified as a key gene for the diagnosis and early detection of open neural tube defects (Karthik et al., 2022). Independently, our mating attempts with p53Y217C/Y217C mice indicated that p53Y217C/Y217C pregnant females exhibited dystocia, and parturition is considered to be a finely regulated inflammatory process (Zhang and Wei, 2021). Of note, dystocia was not observed in p53-/- mice, but was previously reported in pregnant females with a combined deficiency of p53 and FasL (Embree-Ku and Boekelheide, 2002). The fact that p53Y217C/Y217C females shared phenotypic traits with both Bim+/- p53-/- mice (neural tube defects) and FasL-/- p53-/- mice (dystocia) may seem relevant, because Bim and FasL are both regulators of apoptosis and autophagy that also regulate immune responses (Sionov et al., 2015; Taylor and Ng, 2018). Altogether, these data raise the possibility that inflammation may cause accelerated tumorigenesis in p53Y217C/Y217C male mice and might also contribute to promote neural tube defects in p53Y217C/Y217C female embryos or dystocia in p53Y217C/Y217C pregnant females.

In conclusion, we generated a mouse model expressing p53Y217C, the murine equivalent of human p53Y220C, and its analysis revealed that the gain-of-function (or separation-of-function) of a hotspot mutant p53 may not be limited to oncogenic effects. This is an important notion to consider given the ever-expanding functions ascribed to p53 (Hu, 2009; Rakotopare and Toledo, 2023; Voskarides and Giannopoulou, 2023; Vousden and Lane, 2007).

Materials and methods

LSL-Y217C construct

We used mouse genomic Trp53 DNA from previous constructs, containing a portion of intron 1 with a LoxP-Stop-LoxP (LSL) cassette (Ventura et al., 2007) in which the puromycin-resistance gene was replaced by a neomycin-resistance gene (Simeonova et al., 2013). The point mutation (A to G) encoding a Tyr to Cys substitution (TAT to TGT) at codon 217, together with a silent mutation nearby creating a Ban II restriction site, were introduced by PCR directed mutagenesis by using primers Y217C-F and Y217C-R (See Table S6 for primer sequences). The resulting p53LSL-Y217C targeting vector was fully sequenced before use.

Targeting in ES Cells and Genotyping

129/SvJ embryonic stem (ES) cells were electroporated with the targeting construct linearized with Not I. Two recombinant clones, identified by long-range PCR and confirmed by internal PCR and Southern Blot, were injected into C57BL/6J blastocysts to generate chimeras, and germline transmission was verified by genotyping their offspring. In vivo excision of the LSL cassette was performed by breeding p53+/LSL-Y217C male mice with females carrying the PGK-Cre transgene (Lallemand et al., 1998) and genotyping their offspring. The p53Y217C (p53YC) mutation was routinely genotyped by PCR using primers c and d (Figure 1, Table S6) followed by BanII digestion. Tumorigenesis studies were performed on mouse cohorts resulting from ≥ 5 generations of backcrosses with C57BL6/J mice. All experiments were performed according to IACUC regulations as supervised by the Ethics Committee of Institut Curie.

Cells and Cell Culture Reagents

Mouse embryonic fibroblasts (MEFs) isolated from 13.5-day embryos were cultured in a 5% CO2 and 3% O2 incubator, in DMEM GlutaMAX (Gibco), with 15% FBS (Biowest), 100 mM 2- mercaptoethanol (Millipore), 0.01 mM Non-Essential Amino-Acids, and penicillin/streptomycin (Gibco) for ≤ 6 passages. Cells were irradiated with a Cs γ-irradiator or treated for 24 hr with 10 mM Nutlin 3a (Sigma-Aldrich) or 10 mM Doxorubicin (Sigma- Aldrich). Neurospheres were prepared by isolating neural stem/progenitor cells (NPCs) from the lateral and median ganglionic eminences of E14.5 embryos. NPCs were cultured in a 5% CO2 incubator, in DMEM/F12 medium (with L-Glutamine, Invitrogen) supplemented with 1% B-27 (Invitrogen), 1% N2 (Invitrogen), 0.3% glucose (Invitrogen), 25 μg/ml insulin (Sigma-Aldrich), 20 ng/ml EGF (PeproTech), 10 ng/ml bFGF (PeproTech) and penicillin/streptomycin (Gibco). Neurospheres were dissociated once a week by Accutase treatment and mechanically, then seeded at 106 cells per 6 cm diameter dish.

Quantitative RT-PCR

Total RNA, extracted using NucleoSpin RNA II (Macherey-Nagel), was reverse transcribed using SuperScript IV (Invitrogen). Real-time quantitative PCRs were performed as previously described (Simeonova et al., 2012), on an ABI PRISM 7500 using Power SYBR Green (Applied Biosystems). Primer sequences are reported in Table S6.

Western Blots

Cells were lysed in RIPA buffer (50 mM Tris-HCl pH 8, 150 mM NaCl, 5 mM EDTA, 0.5% deoxycholic acid, 0.1% SDS, 1% NP-40) supplemented with a Protease inhibitors cocktail (Roche) and 1 mM PMSF (Sigma). Whole-cell extracts were sonicated three times for 10 s and centrifuged at 13 000 rpm for 30 min to remove cell debris. Protein lysate concentration was determined by BCA assay (Thermo Scientific) and 30 μg of each lysate was fractionated by SDS–PAGE on a 4-12% polyacrylamide gel and transferred onto PDVF membranes (Amersham). Membranes were incubated with antibodies raised against Mdm2 (MDM2-4B2, Abcam, 1/500), p53 (CM-5, Novocastra, 1/2000), p21 (F5, Santa Cruz Biotechnology, 1/200) and actin (actin-HRP sc47778, Santa Cruz Biotechnology, 1/5000). Chemiluminescence revelation of western blots was achieved with the SuperSignal West Dura (Perbio).

ChIP Assay

ChIP analysis was performed as previously described (Simeonova et al., 2013). p53-DNA complexes were immunoprecipitated from total extracts by using 50 μg of a polyclonal antibody against p53 (FL-393, Santa Cruz Biotechnology) and 300 μg of sonicated chromatin. Rabbit IgG (Abcam) was used for control precipitation. Quantitative PCR was performed on ABI PRISM 7500. Primer sequences are reported in Table S6.

Cell Fractionation Assay

Cells were lysed in hypotonic lysis buffer (10 mM Tris pH 7.5, 10 mM NaCl, 3 mM Mgcl2, 0.3% NP-40, 10% glycerol) on ice for 10 min, then centrifuged at 4°C for 3 min at 1000g. The supernatant was recovered as the cytoplasmic fraction, and the pellet was incubated in a modified Wuarin-Schibler buffer (10 mM Tris-HCl pH 7.0, 4 mM EDTA, 0.3 M NaCl, 1 M urea, 1% NP-40) on ice for 5 min, vortexed, incubated on ice for 10 min, then centrifuged at 4°C for 3 min at 1000g. The supernatant was recovered as the nucleoplasmic fraction, and the pellet was washed, then incubated in a nuclear lysis buffer (20 mM Tris pH 7.5, 150 mM KCl, 3 mM MgCl2, 0.3% NP-40, 10% glycerol) on ice for 5 min, sonicated, then centrifuged at 13000 rpm at 4°C and the supernatant was recovered as the chromatin fraction. All buffers were supplemented with a Protease inhibitors cocktail (Roche). Cellular fractions were analyzed by western blots with antibodies against p53 (AF-1355, R&D systems, 1/600), Tubulin (ab15568, Abcam, 1/1000), Nup98 (ab50610, Abcam, 1/1000) and histone H3 (ab1791, Abcam, 1/1000). Chemiluminescence revelation was achieved with the SuperSignal West Pico or Femto (Thermoscientific) and analyzed with a ChemiDoc imaging system (Bio-Rad).

Immunofluorescence

MEFs were cultured on collagen-coated coverslips, treated with 10 μM Nutlin 3a (Sigma-Aldrich) for 24 hr and analyzed. Coverslips were stained with rabbit anti-p53 FL-393 (Santa Cruz Biotechnology), mouse anti-actin primary antibody and with Alexa Fluor 647 anti-Rabbit and Alexa Fluor 488 anti-Mouse secondary antibodies (Molecular Probes). DNA was counterstained with DAPI. Images were captured on an epifluorescence microscope using equal exposure times for all images for each fluor.

Cell-Cycle Assay

Log phase cells were irradiated at room temperature (RT) with a Cs γ-irradiator at doses of 3 or 12 Gy, incubated for 24 hr, then pulse labeled for 1 hr with BrdU (10 μM), fixed in 70% ethanol, double stained with FITC anti-BrdU and propidium iodide, and analyzed by flow cytometry with a BD Biosciences FACSort and the FlowJo software.

Apoptosis Assay

Six to eight weeks-old mice were left untreated or submitted to 10 Gy whole-body ɣ-irradiation. Mice were sacrificed 4 hr later and thymi were extracted. Thymocytes were recovered by filtration through a 70 μm smash, stained with Annexin V-FITC Apoptosis detection kit (Abcam) and propidium iodide, then analyzed with a LSRII FACS machine. Data were analyzed using Flowjo software.

Histology

Organs were fixed in formol 4% for 24 hr, then ethanol 70%, and embedded in paraffin wax. Serial sections were stained as described (Simeonova et al., 2013), with hematoxylin and eosin using standard procedures.

RNA-Seq analysis

Total RNA was extracted from the thymi of 8 weeks-old asymptomatic male mice using nucleospin RNA II (Macherey-Nagel). The quality of RNA was checked with Bioanalyzer Agilent 2100 and RNAs with a RIN (RNA integrity number) ≥ 7 were retained for further analysis. RNA was depleted from ribosomal RNA, then converted into cDNA libraries using a TruSeq Stranded Total Library preparation kit (Illumina). Paired-end sequencing was performed on an Illumina MiSeq platform. Reads were mapped to the mouse genome version GRCm38 and counted on gene annotation gencode.vM18 with featureCounts (Liao et al., 2014). Differentially expressed genes with an adjusted p-value < 0.05 were identified using the DESeq2 R package (Love et al., 2014).

Gene Ontology analysis

Gene ontology analysis of differentially expressed genes was performed by using the GOrilla (Technion) software as previously described (Rakotopare et al., 2023). Enrichment analyses were carried out by comparing the list of 192 differentially expressed genes between p53Y217C/Y217C and p53-/- thymi to the full list of genes (background), with ontology searches for biological processes and default P-value settings (10−3).

Gene set enrichment analysis

Gene set enrichment analysis was performed by using the GSEA software with canonical pathway gene sets from the Mouse Molecular Signature Database (MSigDB) (Subramanian et al., 2005). Gene sets that displayed a normalized enrichment score (NES) > 2, a nominal (NOM) P-value <0.01, and a false discovery rate (FDR) q-value < 0.01 were regarded as significantly enriched.

Statistical analyses

The Student’s t test was used in all figures to analyze differences between two groups of values. For the proportions of mice at weaning, the observed mice count was compared to the expected count according to Mendel’s distribution and a Chi-square (χ2) test, and a Fischer’s test was applied to compare frequencies of mutant females. Survival curves were analyzed with log-rank Mantel-Cox tests. Analyses were performed by using GraphPad Prism, and values of P ≤ 0.05 were considered significant.

eLife Assessment

This work is of fundamental significance and has a compelling level of evidence for the role of mutant p53 in regulation of tumorigenesis using an in vivo mouse model. The study is well-conducted and will be of interest to a broad audience including those interested in p53, transcription factors and cancer biology.

Reviewer #1 (Public review):

Summary:

This manuscript by Toledo and colleagues describes the generation and characterization of Y220C mice (Y217C in the mouse allele). The authors make notable findings: Y217C mice that have been backcrossed to C57Bl/6 for five generations show decreased female pup births due to exencephaly, a known defect in p53 -/- mice, and they show a correlation with decreased Xist expression, as well increased female neonatal death. They also noted similar tumor formation in Y217C/+ and p53 +/- mice, suggesting that Y217C may not function as a dominant negative. Notably, the authors find that homozygous Y217C mice die faster than p53 -/- mice and that the lymphomas in the Y217C mice were more aggressive and invasive. The authors then perform RNA seq on thymi of Y217C homozygotes compared to p53 -/-, and they suggest that these differentially expressed genes may explain the increased tumorigenesis in Y217C mice.

Strengths:

Overall, the study is well controlled and quite well done and will be of interest to a broad audience, particularly given the high frequency of the Y220C mutation in cancer (1% of all cancers, 4% of ovarian cancer).

Weaknesses:

No weaknesses were noted by this reviewer.

Reviewer #2 (Public review):

Summary:

Jaber et al. describe the generation and characterization of a knock-in mouse strain expressing the p53 Y217C hot-spot mutation. While the homozygous mutant cells and mice reflect the typical loss-of-p53 functions, as expected, the Y217C mutation also appears to display gain-of-function (GOF) properties, exemplified by elevated metastasis in the homozygous context (as noted with several hot-spot mutations). Interestingly, this mutation does not appear to exhibit any dominant-negative effects associated with most hot-spot p53 mutations, as determined by the absence of differences in overall survival and tumor predisposition of the heterozygous mice, as well as target gene activation upon nutlin treatment.

In addition, the authors noted a severe reduction in the female 217/217 homozygous progeny, significantly more than that observed with the p53 null mice, due to exencephaly, leading them to conclude that the Y217C mutation also has additional, non-cancer-related GOFs. Though this property has been well described and attributed to p53 functional impairment, the authors conclude that the Y217C has additional properties in accelerating the phenotype.

Transcriptomic analyses of thymi found additional gene signature differences between the p53 null and the Y217C strains, indicative of novel target gene activation, associated with inflammation.

Strengths:

Overall, the characterisation of the mice highlights the expected typical outcomes associated with most hot-spot p53 mutations published earlier. The quality of the work presented is well done and good, and the conclusions and reasonably well justified.

Weaknesses:

The manuscript would benefit from the provision of additional data to strengthen the claims made, as follows:

(1) Oncogenic GOF - the main data shown for GOF are the survival curve and enhanced metastasis. Often, GOF is exemplified at the cellular level as enhanced migration and invasion, which are standard assays to support the GOF. As such, the authors should perform these assays using either tumor cells derived from the mice or transformed fibroblasts from these mice. This will provide important and confirmatory evidence for GOF for Y217C.

(2) Novel target gene activation - while a set of novel targets appears to be increased in the Y217C cells compared to the p53 null cells, it is unclear how they are induced. The authors should examine if mutant p53 can bind to their promoters through CHIP assays, and, if these targets are specific to Y217C and not the other hot-spot mutations. This will strengthen the validity of the Y217C's ability to promote GOF.

(3) Dominant negative effect - the authors' claim of lack of DN effect needs to be strengthened further, as most p53 hot-spot mutations do exhibit DN effect. At the minimum, the authors should perform additional treatment with nutlin and gamma irradiation (or cytotoxic/damaging agents) and examine a set of canonical p53 target genes by qRT-PCR to strengthen their claim.

Author response:

The following is the authors’ response to the original reviews

Reviewer #2 suggested the addition of new data to address the following points:

Reviewer #2:

(1) Oncogenic GOF - the main data shown for GOF are the survival curve and enhanced metastasis. Often, GOF is exemplified at the cellular level as enhanced migration and invasion, which are standard assays to support the GOF. As such, the authors should perform these assays using either tumor cells derived from the mice or transformed fibroblasts from these mice. This will provide important and confirmatory evidence for GOF for Y217C.

We thank the referee for this comment. Our previous data indicated accelerated tumor progression and increased metastasis in Trp53Y217C/Y217C mice, which provided in vivo evidence of an oncogenic gain of function (GOF) for the p53Y217C mutant. However, we agree that it was important to provide additional evidence of GOF at the cellular level.

Many cellular assays were previously used to evaluate the GOF of p53 mutants, including those listed by the referee. Importantly, Zhao et al. recently showed that a common property of several p53 mutants proposed to have oncogenic GOF is their capacity to promote chromosomal instability (Zhao et al. (2024) Nat. Commun. 15, 180). For the revision of our manuscript, we compared the frequencies of chromosomal alterations occurring spontaneously in WT, Trp53Y217C/Y217C and Trp53-/- mouse embryonic fibroblasts (MEFs). Chromosome breaks, radial chromosomes and DMs were more frequent in Trp53Y217C/Y217C MEFs than in WT or Trp53-/- MEFs, providing clear evidence of a GOF promoting chromosomal instability. This new result is presented in Figure 2G and mentioned in the revised abstract.

Furthermore, as pointed out by referee #1 in a confidential comment, increased NF-kB signaling provides evidence of p53 GOF. Accordingly, Zhao et al. proposed that the capacity of p53G245D and p53R273H to promote chromosomal instability ultimately led to activation of a noncanonical NF-kB signaling that would promote tumor cell invasion and metastasis. Consistent with their work, we now report that the GSEA of Trp53Y217C/Y217C and Trp53-/- thymocytes revealed an upregulation of non-canonical NF-kB signaling in Trp53Y217C/Y217C thymic cells (a new result presented in Figure 5F and Supplementary Figure S13). These new data lead us to mention in the revised discussion that “similar mechanisms might underlie the oncogenic properties of the p53Y217C, p53G245D and p53R273H mutants”.

(2) Novel target gene activation - while a set of novel targets appears to be increased in the Y217C cells compared to the p53 null cells, it is unclear how they are induced. The authors should examine if mutant p53 can bind to their promoters through CHIP assays, and, if these targets are specific to Y217C and not the other hot-spot mutations. This will strengthen the validity of the Y217C's ability to promote GOF.

We respectfully disagree with the referee when he/she considers that the validity of p53Y217C’s ability to promote a GOF would be strengthened by showing that p53Y217C binds to the promoters of genes upregulated in Trp53Y217C/Y217C cells. In fact, Pal et al. recently performed the experiment proposed by the referee, by integrating RNAseq and ChIPseq data from MCF10A cells expressing p53Y220C, the human equivalent of p53Y217C, and found that 95% of the genes upregulated upon p53Y220C expression were upregulated indirectly, without p53Y220C binding to their promoters (Pal et al. (2023) NPJ Breast Cancer 9, 78). Consistent with our data, Pal et al. notably found that the expression of p53Y220C increased cell migration and invasion, which correlated with an increased expression of S100A8 and S100A9. They found that the promoters of S100A8 and S100A9 were however not bound by p53Y220C, indicating an indirect mechanism for their upregulated expression. Furthermore, the study by Zhao et al. mentioned above also suggested an indirect mechanism of GOF, because the upregulation of inflammation-related genes by a mutant p53 protein was proposed to result from signaling cascades triggered by chromosomal instability. Our data appear consistent with both studies, because p53Y217C was undetectable or barely detectable in the chromatin fraction of Trp53Y217C/Y217C cells, and because Trp53Y217C/Y217C cells exhibited increased chromosome instability and increased NFB signaling compared to Trp53-/- cells, which may suggest indirect mechanisms for p53Y217C GOF.

Nevertheless, we agree with the referee that it was important to provide stronger evidence of p53Y217C GOF in the revised manuscript. In that regard, we were intrigued by the perinatal death of most Trp53Y217C/Y217C females, which provided evidence of unexpected teratogenic effects of the mutant. We had proposed that these female-specific teratogenic effects likely resulted from pro-inflammatory GOF of p53Y217C. This hypothesis relied on the RNAseq pro-inflammatory signature in Trp53Y217C/Y217C thymic cells, and on the fact that the glycoprotein CD44, known to drive inflammation, had been identified as a key gene in open neural tube defects. However, we had not tested this hypothesis experimentally. In the revised version of the manuscript, we tested this hypothesis. We mated Trp53+/Y217C female mice with Trp53Y217C/Y217C males, then administered supformin (LCC-12), a potent CD44 inhibitor known to attenuate inflammation in vivo, to pregnant mice by oral gavage. The administration of subformin led to a five-fold increase in the proportion of weaned Trp53Y217C/Y217C females in the progeny, suggesting that reducing inflammation in utero rescued some of the Trp53Y217C/Y217C female embryos. This new result is presented in Figure 5G and Supplementary Table S6, and mentioned in the abstract.

We believe that these new results, as well as the additional GSEA analyses revealing increased NFkB signaling in Trp53Y217C/Y217C cells, further emphasize the importance of inflammation in the GOF of the p53Y217C mutant. Accordingly, we slightly modified the title of our article, to include the notion that Trp53Y217C is an inflammation-prone mouse model. We also end the article by summarizing the effects of p53Y217C in vivo, in a new Supplementary Table S7 that compares the LOF effects of a p53 KO with the (LOF+GOF) effects of the p53Y217C mutant.

(3) Dominant negative effect - the authors' claim of lack of DN effect needs to be strengthened further, as most p53 hot-spot mutations do exhibit DN effect. At the minimum, the authors should perform additional treatment with nutlin and gamma irradiation (or cytotoxic/damaging agents) and examine a set of canonical p53 target genes by qRT-PCR to strengthen their claim.

Our previous data indicated identical tumor onset and survival in Trp53+/Y217C and Trp53+/- mice, leading us to conclude that, at least for spontaneous tumorigenesis, there was no evidence of a Dominant Negative Effect (DNE) in vivo. Here, we followed the referee’s suggestion and evaluated the possibility of a DNE in response to stress, by comparing WT, Trp53+/Y217C and Trp53+/- MEFs or thymocytes. We analyzed different types of stress (Nutlin, Doxorubicin, girradiation) and different types of cellular responses (transactivation of classical p53 target genes, cell cycle arrest, apoptosis), and the results lead us to conclude that there is little if any DNE also in response to various stresses. These new data are mentioned in a paragraph evaluating the possibility of DNE or GOF at the cellular level, and presented in a new Supplementary Figure S6.

Version 2

Abstract

Abstract

Missense “hotspot” mutations localized in six p53 codons account for 20% of TP53 mutations in human cancers. Hotspot p53 mutants have lost the tumor suppressive functions of the wildtype protein, but whether and how they may gain additional functions promoting tumorigenesis remain controversial. Here we generated Trp53Y217C, a mouse model of the human hotspot mutant TP53Y220C. DNA damage responses were lost in Trp53Y217C/Y217C cells, and Trp53Y217C/Y217C fibroblasts exhibited increased chromosome instability compared to Trp53-/- cells. Furthermore, Trp53Y217C/Y217C male mice died earlier than Trp53-/- males, with more aggressive thymic lymphomas. This correlated with an increased expression of inflammation-related genes in Trp53Y217C/Y217C thymic cells compared to Trp53-/- cells. Surprisingly, we recovered only one Trp53Y217C/Y217C female for 22 Trp53Y217C/Y217C males at weaning, a skewed distribution explained by a high frequency of Trp53Y217C/Y217C female embryos with exencephaly and the death of most Trp53Y217C/Y217C female neonates. Strikingly however, when we treated pregnant females with the anti-inflammatory drug supformin (LCC-12) we observed a five-fold increase in the proportion of viable Trp53Y217C/Y217C weaned females in their progeny. Together, these data suggest that the p53Y217C mutation not only abrogates wildtype p53 functions but also promotes inflammation, with oncogenic effects in males and teratogenic effects in females.

Introduction

Somatic alterations in the TP53 gene, encoding the tumor suppressor p53, are the most common events in human tumors (Hollstein et al., 1991). The p53 protein is a stress sensor stabilized and activated in response to potentially oncogenic insults. In its tetrameric active form, wildtype (WT) p53 can trigger a transcriptional program to induce various responses including cell cycle arrest, senescence, apoptosis or metabolic changes (Beckerman and Prives, 2010). In human cancers, about 75% of all TP53 alterations are missense mutations, and the eight most frequent (“hotspot”) missense mutations (R175H, Y220C, G245S, R248Q, R248W, R273H, R273C and R282W) affect six residues localized in the p53 DNA-binding domain (DBD) and account for 20% of all TP53 mutations (Hainaut and Pfeifer, 2016). The fact that most TP53 mutations are missense substitutions suggested that cells expressing mutant p53 might have a selective advantage over cells lacking p53, and evidence for this was first gained by expressing various p53 mutants in p53-null cells and observing enhanced tumorigenic potential in nude mice or increased plating efficiency in agar cell culture (Dittmer et al., 1993). Mouse models expressing hotspot p53 mutants next helped to define potential mechanisms accounting for accelerated tumorigenesis. First, a dominant-negative effect (DNE) may be observed in heterozygotes, if the mutant p53 inhibits the WT p53 protein in hetero-tetramers (Gencel-Augusto and Lozano, 2020). Evidence that this mechanism accounts for accelerated tumorigenesis was notably reported in human leukemia (Boettcher et al., 2019). A second mechanism is a gain-of-function (GOF), i.e. the acquisition by mutant p53 of new oncogenic properties (Amelio and Melino, 2020; Stein et al., 2020). Although the DNA sequence specificity of p53 mutants is impaired, leading to a loss of WT functions (LOF), many p53 mutant proteins are stabilized in the cell and might engage in aberrant interactions with other transcription factors, chromatin-modifying complexes or DNA helicase subunits, leading to the acquisition of GOF phenotypes (Kim and Deppert, 2007; Pfister and Prives, 2017; Zhao et al., 2024). However, a few recent studies challenged the pathological importance of mutant p53 GOF, or at least its relevance in anti-cancer therapeutic strategies (Aubrey et al., 2018; Boettcher et al., 2019; Wang et al., 2023). An alternative hypothesis emerged, postulating that the tumorigenic properties of a p53 mutant might result from a separation-of-function (SOF), if the mutant retains pro-proliferative or pro-survival functions of WT p53 while losing its tumor suppressive activities (Kennedy and Lowe, 2022). The concept of SOF was first proposed for TP53 exon-6 truncating mutations, which mimic a naturally occurring alternative p53 splice variant (Shirole et al., 2016), but may apply to some p53 missense mutations, for example p53R248W (Humpton et al., 2018). Importantly, the analysis of various mutant p53 alleles in vivo appears crucial for a better understanding of their contribution to tumorigenesis, because GOF/SOF phenotypes may vary depending on the mutated residue, its specific mutation, the cellular context or genetic background (Dibra et al., 2024; Hanel et al., 2013; Kadosh et al., 2020; Kim and Lozano, 2018; Kotler et al., 2018; McCann et al., 2022; Xiong et al., 2022).

In human cancers, the TP53Y220C mutation is the most frequent missense mutation outside of the DNA-binding surface of p53 (Hainaut and Pfeifer, 2016). The somatic TP53Y220C mutation accounts for over 100,000 new cancer cases per year worldwide, including solid tumors and myeloid neoplasms, and a germline TP53Y220C mutation was reported in 15 families with the Li-Fraumeni syndrome (LFS) of cancer predisposition (Bouaoun et al., 2016; Gener-Ricos et al., 2024). The mutation causes a structural change in the DBD localized at the periphery of the β-sandwich region of the protein, far from the surface contact of DNA. The mutation from a tyrosine to a cysteine markedly lowers the thermodynamic stability of the DBD, leading to a largely unfolded and inactive protein at body temperature (Joerger et al., 2006), and a molecule designed to bind p53Y220C and restore WT protein conformation appears as a promising anti-cancer drug (Dumbrava et al., 2022). Analyses of the impact of p53Y220C in cancer cell lines led to conflicting results. On one hand, overexpression of p53Y220C in a p53-null cell line increased its capacity for migration or invasion (Zhou et al., 2022), and a decreased expression of p53Y220C, caused by RNA interference or various chemical compounds, promoted cell death or decreased the migratory or invasive capacities of cells (Tseng et al., 2022; Vikhanskaya et al., 2007; Zhou et al., 2022). On the other hand, the removal by CRISPR/Cas9 of p53 mutants with reported GOF - including p53Y220C - in diverse cancer cell lines had no significant impact on cell survival or proliferation, and the cell death caused by RNA interference against p53Y220C might result from non-specific toxic effects (Wang et al., 2023). Here, to gain a better understanding of the impact of the p53Y220C mutation in vivo, we generated a mouse model with a targeted Trp53Y217C mutation - the mouse equivalent to human TP53Y220C (Figure S1) - and analyzed animals and cells carrying one or two copies of the mutant allele.

Results

Targeting of a p53Y217C mutation in the mouse

We used homologous recombination in 129/SvJ ES cells to target the p53Y217C mutation at the mouse Trp53 locus. Mice expressing p53Y217C conditionally were generated by using a targeting vector containing transcriptional stops flanked by LoxP sites (LSL) upstream of coding sequences, and the p53Y217C mutation in exon 6 (Figure 1A-D). Trp53+/LSL-Y217C mice were then crossed with mice carrying the PGK-Cre transgene (Lallemand et al., 1998) to excise the LSL cassette and obtain Trp53+/Y217C mice, expressing the mutant protein constitutively. After two backcrosses over a C57BL/6J background, we prepared mouse embryonic fibroblasts (MEFs) from intercrosses of Trp53+/Y217C mice (Figure 1E). We extracted RNAs from Trp53Y217C/Y217C MEFs then sequenced p53 mRNAs to verify that the p53Y217C coding sequence was identical to the wildtype p53 (p53WT) sequence, except for the desired missense mutation and a silent mutation introduced to facilitate mouse genotyping (Figure 1F). The quantification of p53 mRNA levels from WT and Trp53Y217C/Y217C MEFs next revealed similar transcription from both alleles (Figure 1G).

Targeting the Y217C missense mutation at the mouse Trp53 locus.

(A) Targeting strategy. The wildtype (WT) Trp53 gene is within a 17 kb-long EcoRI (RI) fragment (black boxes are for coding sequences and white boxes for UTRs). The targeting construct contains: (1) a 1.5 kb-long 5’ homology region; (2) a Lox-Stop-Lox (LSL) cassette with a neomycin selection gene (Neo), four transcriptional stops (STOP) and an EcoRI site, flanked by LoxP sites (arrowheads); (3) p53 coding exons, including the Y217C (YC) missense mutation in exon 6 (asterisk) and an additional BanII site; (4) a 2.8 kb-long 3’ homology region; and (5) the diphteria α-toxin (DTA) gene for targeting enrichment. Proper recombinants with a Trp53LSL-Y217C allele, resulting from the described crossing-overs, were G418 resistant. They were identified by a 2.4 kb-long band after PCR with primers a and b, and confirmed by bands of 635 and 224 bp after PCR with primers c and d and BanII digestion. They were also verified by Southern blot with the indicated probe as containing a 10.5 kb EcoRI band. Two recombinant ES clones were injected into blastocysts to generate chimeras, and germline transmission was verified by genotyping with primers c and d and BanII digestion. Excision of the LSL cassette was performed in vivo, by breeding Trp53+/LSL-Y217C male mice with females carrying the PGK-Cre transgene, to obtain mice with a Trp53Y217C allele. (B-D) Screening of recombinant ES clones (+) by PCR with primers a and b (B); PCR with primers c and d then BanII digestion (C); Southern blot (D). (E) Genotyping of mouse embryonic fibroblasts (MEFs) from an intercross of Trp53+/Y217C mice, by PCR with primers c and d and BanII digestion. (F) Trp53Y217C sequence around codon 217. The introduced Y217C missense mutation and the silent mutation creating an additional BanII restriction site are highlighted (asterisks). (G) WT and Trp53Y217C/Y217C (YC/YC) MEFs express similar p53 mRNA levels. Total RNA was extracted, then p53 mRNAs were quantified by real-time qPCR, normalized to control mRNAs and the amount in WT cells was assigned the value of 1. Primer sequences are listed in Table S8.

Deficient p53-dependent stress responses in Trp53Y217C/Y217C cells

We used western blots to analyze protein extracts from Trp53+/+, Trp53+/-, Trp53+/Y217C, Trp53Y217C/Y217C and Trp53-/- MEFs, unstressed or treated with Nutlin, an antagonist of Mdm2, the E3 ubiquitin ligase for p53 (Vassilev et al., 2004). Results indicated a high increase in p53 abundance in untreated and Nutlin-treated Trp53Y217C/Y217C MEFs and a moderate increase in untreated and Nutlin-treated Trp53+/Y217C MEFs, compared to WT cells (Figures 2A and S2). Protein levels for p21 and Mdm2, the products of two classical p53 target genes, appeared similar in Trp53+/+ and Trp53+/Y217C MEFs, and were undetectable or markedly decreased in Trp53Y217C/Y217C and Trp53-/- MEFs (Figures 2A and S2). Accordingly, p53Y217C appeared unable to transactivate the Cdkn1a/p21 and Mdm2 genes or to bind their promoters (Figure 2B). The fractionation of cells before or after treatment with doxorubicin, a clastogenic drug, indicated that p53WT accumulated in the nucleoplasm and chromatin fractions in response to DNA damage, whereas p53Y217C appeared mostly cytoplasmic in both untreated and doxorubicin-treated cells, and undetectable or barely detectable in chromatin fractions (Figures 2C and S3). Furthermore, when Nutlin-treated MEFs were analyzed by immunofluorescence, p53WT was only detected in nuclei, whereas p53Y217C could be observed in nuclear and cytoplasmic compartments (Figure 2D). We next analyzed two well-known p53-mediated responses to DNA damage, i.e. cell cycle arrest in MEFs and apoptosis in thymocytes (Kastan et al., 1992; Lowe et al., 1993). WT MEFs exposed to increasing doses of ɣ-irradiation (3 Gy or 12 Gy) exhibited significant increases in G1/S ratios, whereas G1/S ratios were similar before or after irradiation in Trp53Y217C/Y217C and Trp53-/- MEFs (Figures 2E and S4). Furthermore, thymocytes recovered from irradiated WT mice underwent a massive apoptotic response, with an almost three-fold increase in apoptotic cells after a 10 Gy irradiation. By contrast, no increase in apoptotic cells was observable upon irradiation in the thymi from Trp53Y217C/Y217C mice, and apoptotic thymocytes were equally rare in irradiated Trp53Y217C/Y217C and Trp53-/- mice (Figures 2F and S5). Together, these results indicated that the p53Y217C mutation altered the abundance and intracellular distribution of the p53 protein, associated with a decrease in DNA-bound protein, and that p53Y217C had completely lost the ability to induce cell cycle arrest and apoptosis upon DNA damage.

Trp53Y217C/Y217C cells exhibit alterations in p53 abundance and subcellular localization, defective responses to DNA damage, and increased chromosomal instability.

(A) Increased p53 protein levels in Trp53Y217C/Y217C and Trp53+/Y217C MEFs. MEFs of the indicated genotypes were treated or not with 10 μM Nutlin 3a for 24 h, then protein extracts were immunoblotted with antibodies against Mdm2, p53, p21 and actin. (B) The transactivation of classical p53 target genes Cdkn1a/p21 and Mdm2 is impaired in Trp53Y217C/Y217C cells. WT, Trp53Y217C/Y217C and Trp53-/- MEFs were treated as in (A), then (top) mRNAs were quantified in 5-6 independent experiments using real-time PCR, with results normalized to control mRNAs and mean RNA amounts in unstressed WT cells assigned a value of 1; or (bottom) ChIP assays were performed at the Cdkn1a and Mdm2 promoters in 2-3 independent experiments with an antibody against p53 or rabbit IgG as a negative control. Immunoprecipitates were quantified using real-time PCR, normalized to data over an irrelevant region, and the amount in unstressed WT cells was assigned a value of 1. (C) Assessment of p53WT and p53Y217C subcellular localization by cellular fractionation. WT and Trp53Y217C/Y217C MEFs were treated or not with 1 μM doxorubicin (Doxo) for 24 hr, submitted to cellular fractionation, then protein extracts were immunoblotted with antibodies against p53 or the fraction controls Tubulin for cytoplasm (Cp.), Nup98 for nucleoplasm (Np.) and histone H3 for chromatin (χin). (D) Assessment of p53WT and p53Y217C subcellular localization by immunofluorescence. WT, Trp53Y217C/Y217C and Trp53-/- MEFs were treated with 10 μM Nutlin 3a for 24 h, then stained with antibodies against p53 (red) or actin (green) and DNA was counterstained with DAPI (blue). (E) Absence of a cell cycle arrest response in Trp53Y217C/Y217C MEFs. Asynchronous cell populations of Trp53+/+, Trp53Y217C/Y217C and Trp53-/- MEFs were analyzed 24 hr after 0, 3 or 12 Gy ɣ-irradiation. Results from 3 independent experiments. (F) Absence of a p53-dependent apoptotic response in Trp53Y217C/Y217C thymocytes. Age-matched mice of the indicated genotypes were left untreated or submitted to 10 Gy whole-body ɣ-irradiation then sacrificed after 4 hr and their thymocytes were stained with Annexin V-FITC and analyzed by FACS, in 2 independent experiments. (G) Increased chromosomal instability in Trp53Y217C/Y217C fibroblasts. Metaphase spreads were prepared from WT, Trp53Y217C/Y217C and Trp53-/- MEFs at passage 4, then aberrant metaphases (with chromosome breaks, radial chromosomes or DMs) were scored. Left: distribution of aberrant metaphases. Data from 110 WT, 97 Trp53Y217C/Y217C or 119 Trp53-/- complete diploid metaphases, independently observed by two experimenters. Right: examples of two aberrant Trp53Y217C/Y217C metaphases: one with a double minute chromosome (DM), a chromosome break (Br) and a radial chromosome (R), the other with multiple DMs. Enlargements of regions of interest are presented between the two metaphases. Scale bars (D, G): 5 μm. ***P<0.001, **P<0.01, *P<0.05, °P=0.09, ns: non-significant by Student’s t (B, E, F) or Fischer’s (G) tests.

Increased chromosomal instability in Trp53Y217C/Y217C cells

We next searched for evidence, at the cellular level, of potential dominant negative effect (DNE) or gain of function (GOF) for the p53Y217C mutant. Little if any difference in the transactivation of canonical p53 target genes was observed between WT, Trp53+/Y217C and Trp53+/- cells treated with Nutlin or Doxorubicin (Figure S6A), and Trp53+/Y217C and Trp53+/- cells exhibited similar cell cycle arrest or apoptotic responses to γ-irradiation (Figure S6B-C), suggesting little or no DNE in response to various stresses.

On the opposite, we obtained clear evidence of a GOF for the p53Y217C mutant. Two p53 mutants (p53G245D and p53R273H) were recently proposed to promote tumorigenesis by predisposing cells to chromosomal instability (Zhao et al., 2024), which led us to compare the frequency of chromosome rearrangements in WT, Trp53Y217C/Y217C and Trp53-/- primary MEFs. We searched for chromosome breaks, radial chromosomes, or double-minute chromosomes (DMs) in preparations of MEFs treated with the anti-mitotic nocodazole for metaphase enrichment. The three categories of chromosome rearrangements were more frequently observed in Trp53Y217C/Y217C MEFs than in Trp53-/- or WT cells (Figure 2G), indicating a GOF that promotes chromosomal instability.

Impact of p53Y217C on mouse development

We next determined the impact of the p53Y217C mutation in vivo, by comparing mouse cohorts generated from intercrosses of heterozygous Trp53+/- or Trp53+/Y217C mice resulting from ≥ 5 backcrosses to the C57BL6/J background. Intercrosses of Trp53+/- mice are known to yield a reduced proportion of Trp53-/- mice, that is mainly due to defects in neural tube closure causing exencephaly in a fraction of Trp53-/- female embryos (Armstrong et al., 1995; Sah et al., 1995) and, to a lesser extent, to lung dysfunction affecting a fraction of Trp53-/- neonates (Tateossian et al., 2015). At weaning (on the 21st day postpartum or P21), we observed one Trp53-/- female mouse for 3.5 p53-/- males from Trp53+/- intercrosses, an underrepresentation of females consistent with frequencies reported in earlier studies (Armstrong et al., 1995; Sah et al., 1995). Strikingly, the underrepresentation of weaned females was even more acute for Trp53Y217C/Y217C mice, with only one Trp53Y217C/Y217C female for 19 Trp53Y217C/Y217C males from Trp53+/Y217C intercrosses (Figure 3A). We next analyzed Trp53Y217C/Y217C embryos generated from heterozygous intercrosses or from heterozygous-homozygous (Trp53+/Y217C x Trp53Y217C/Y217C) crosses, at embryonic days E12.5-E16.5. An underrepresentation of Trp53Y217C/Y217C female embryos was not observed, but 11/26 (42%) female embryos exhibited developmental abnormalities, including 10 with exencephaly, whereas all the male embryos appeared normal (Figure 3B). Importantly, the frequency of Trp53Y217C/Y217C female embryos with exencephaly (38.5%) was much higher than the reported frequency (0-8%) of Trp53-/- female embryos with exencephaly of C57BL/6 x 129/Sv genetic background (Donehower et al., 1992; Sah et al., 1995), suggesting a stronger effect of the p53Y217C mutant on female embryonic development.

Trp53Y217C/Y217C mice exhibit female-specific perinatal lethality.

(A) Distribution of weaned mice obtained from Trp53+/- or Trp53+/Y217C intercrosses. Obs: observed numbers of mice at weaning (P21); exp: expected numbers assuming a Mendelian distribution without sex distortion; f/m: observed female/male ratios. Consistent with previous reports, the observed distribution of weaned mice from Trp53+/- intercrosses did not conform to values expected for a Mendelian distribution without sex distortion (υ=5; χ2=16.31>15.09), indicating a significant deficit in female Trp53-/- mice (top). The distribution of weaned mice from Trp53+/Y217C intercrosses diverged even more from values for a Mendelian distribution without sex distortion (υ =5; χ2=104.23>15.09), due to a striking deficit in female Trp53Y217C/Y217C mice (bottom). Differences between the frequencies of Trp53Y217C/Y217C (4/677) and Trp53-/- (8/196) females in the progeny, or between the female to male ratios for Trp53Y217C/Y217C (4/75) and Trp53-/- (8/28) animals, are statistically significant (P=0.0012 and P=0.0087 in Fischer’s tests, respectively). (B) Exencephaly is frequently observed in p53Y217C/Y217C female embryos. Top: the distribution of E12.5-16.5 embryos from heterozygous (Trp53+/Y217C) intercrosses or heterozygous-homozygous (Trp53+/Y217C x Trp53Y217C/Y217C) crosses is shown. f or m exenc.: number of female or male embryos with exencephaly; o.a.: embryos with other abnormalities. Below, examples of female Trp53Y217C/Y217C embryos at E12.5, 13.5 and 16.5 exhibiting exencephaly (arrows) are each shown (center) together with a normal embryo from the same litter (bottom). (C) Distribution of mice at birth from the indicated crosses. Of note, out of five Trp53Y217C/Y217C females observed at birth, only one remained alive at weaning age. Thus, the female/male ratio for weaned Trp53Y217C/Y217C animals from these crosses was 1/22, a ratio similar to the one observed in A (4/75).

In Trp53-/- female mice, exencephaly was previously correlated with stochastic aberrant X chromosome inactivation, with a decreased expression of Xist and an increase in bi-allelic expression of X-linked genes including Maob, Pls3 and Usp9x (Delbridge et al., 2019). We prepared neurospheres from Trp53+/+, Trp53-/- and Trp53Y217C/Y217C female embryos and quantified mRNAs for these genes in neurospheres. Compared to neurospheres from WT animals, Trp53Y217C/Y217C and Trp53-/- neurospheres exhibited an apparent decrease in Xist expression and significantly higher levels of Maob, Pls3 and Usp9x (Figure S7). Thus, as for female Trp53-/- mice, aberrant chromosome X inactivation may contribute to the underrepresentation of Trp53Y217C/Y217C female mice. Of note however, defects in chromosome X inactivation did not appear more pronounced in Trp53Y217C/Y217C neurospheres than in Trp53-/- neurospheres.

To further analyze the impact of the p53Y217C mutation during development, we also determined its potential effect in embryos lacking Mdm2 or Mdm4, two major p53 regulators. The embryonic lethality resulting from Mdm2 or Mdm4 loss is rescued by a concomitant p53 deficiency (Bardot et al., 2015; Finch et al., 2002; Jones et al., 1995; Migliorini et al., 2002; Montes de Oca Luna et al., 1995; Parant et al., 2001). This provides a powerful assay for analyzing the functionality of p53 mutant alleles (Bardot et al., 2015; Iwakuma et al., 2004; Marine et al., 2006; Toledo et al., 2006). We identified Mdm2-/- Trp53Y217C/Y217C and Mdm4-/- Trp53Y217C/Y217C viable mice of both sexes, consistent with a major loss of canonical WT p53 activities in the p53Y217C mutant (Table S1). Of note, in these experiments we mated Mdm4+/- Trp53+/Y217C mice with Mdm4+/- Trp53Y217C/Y217C mice, or Mdm2+/- Trp53+/Y217C mice with either Mdm2-/- Trp53Y217C/Y217C or Mdm2+/- Trp53Y217C/Y217C animals, to analyze the progeny at weaning. No Trp53Y217C/Y217C female mouse was recovered from these crosses, whereas 14 Trp53Y217C /Y217C males were obtained, again demonstrating a striking deficit in Trp53Y217C/Y217C females at weaning. Among animals lacking one or two copies of either Mdm2 or Mdm4, we also observed a deficit in Trp53Y217C/Y217C weaned females. However, the female to male ratio for Trp53Y217C/Y217C animals appeared markedly increased in genetic backgrounds with Mdm4 haploinsufficiency (4/15) or loss (1/6), but only marginally increased (if at all) in genetic backgrounds with Mdm2 haploinsufficiency (1/11) or loss (1/11). This suggested that altering the levels of p53 inhibitors, particularly Mdm4, might improve the survival of Trp53Y217C/Y217C females (Table S1).

The high frequency (42%) of abnormal Trp53Y217C/Y217C female embryos at E12.5-E16.5 could only partially account for the acute deficit in Trp53Y217C/Y217C females observed at weaning, suggesting either embryonic abnormalities undetectable macroscopically, or that a fraction of Trp53Y217C/Y217C females died later in development or postpartum. We performed additional crosses to analyze female to male ratios at postpartum day 0 or 1 (P0-P1) and observed 20 females for 18 males for Trp53+/+ animals, but only five females for 22 males for Trp53Y217C/Y217C animals (Figure 3C). Furthermore, only one of these Trp53Y217C/Y217C females reached weaning age, and the four other neonates died before P2. Altogether, these data led to conclude that the p53Y217C mutation caused female-specific developmental abnormalities at a higher penetrance than a null allele, leading to a perinatal (late embryonic or early postpartum) lethality for most homozygous mutant females.

Interestingly, we observed that Trp53Y217C/Y217C males were fertile and included them in some of our crosses to generate homozygous mutants at higher frequencies (Figure 3B-C). By contrast, we were able to mate two Trp53Y217C/Y217C adult females with a Trp53+/Y217C male, and both got pregnant but had to be euthanized due to complications while giving birth. In both cases, extended labor (> 24 hr) was the main sign of dystocia (Table S2). These observations suggest that, even for the rare Trp53Y217C/Y217C females able to reach adulthood, the p53Y217C mutation caused pathological traits not observed in Trp53-/- females, because the loss of p53 was not previously reported to cause dystocia (Embree-Ku and Boekelheide, 2002; Guimond et al., 1996; Hu et al., 2007). Together, our data indicated that the p53Y217C mutation not only abolished canonical p53 activities, but also conferred additional effects compromising female perinatal viability or parturition.

Impact of p53Y217C on tumorigenesis

We next analyzed the impact of the p53Y217C mutation on the onset and spectrum of spontaneous tumors in mice. We first compared Trp53+/Y217C, Trp53+/- and Trp53+/+ mouse cohorts for 2 years. Trp53+/Y217C and Trp53+/- cohorts exhibited similar spontaneous tumor onset and spectrum, with most mice developing sarcomas during their second year of life (Figure S8). Thus, at least for tumors arising spontaneously, the p53Y217C mutant protein did not appear to exert a DNE over the wildtype p53 protein.

We next compared Trp53Y217C/Y217C and Trp53-/- mouse cohorts. Because of the difficulty to generate Trp53Y217C/Y217C females, we restricted our comparison to males, to avoid potential biases that might result from different sex ratios. All Trp53-/- mice are known to die in less than 1 year, from thymic lymphomas in most cases, or more rarely from sarcomas (Donehower et al., 1992; Jacks et al., 1994). Trp53Y217C/Y217C males died faster than their Trp53-/- counterparts: all the Trp53Y217C/Y217C males were dead by the age of 7 months, whereas more than 20% of the Trp53-/- males were still alive at that age (Figure 4A). Furthermore, most Trp53-/- and Trp53Y217C/Y217C mice died from thymic lymphomas, but histological analysis of tumor organs revealed that the lymphomas in Trp53Y217C/Y217C males were more aggressive and invasive, with sites of metastases notably including the lungs, spleen, liver or kidneys (Figures 4B-C and S9). Thus, the p53Y217C mutation not only abolished canonical tumor suppressive activities but also apparently conferred oncogenic properties to the encoded protein. Consistent with this, when we performed an RNA-seq analysis of thymi from 8 weeks-old Trp53+/+, Trp53-/- and Trp53Y217C/Y217C males, 81.5 % of the 717 differentially expressed genes indicated a loss of function (LOF) in the p53Y217C mutant, but genes suggesting a GOF or a separation of function (SOF) were also observed (Figure S10A and Table S3). Of note, among the differentially expressed genes corresponding to a LOF in the p53Y217C mutant were Bbc3/Puma, Cdkn1a/p21 and Zmat3 (Figure S10B), three p53-target genes known to play major roles in p53-mediated tumor suppression (Brennan et al., 2023).

Oncogenic effects of the mutant protein in Trp53Y217C/Y217C male mice.

(A-B) In homozygous males, p53Y217C leads to accelerated tumor-induced death (A), and aggressive metastatic tumors (B); n=cohort size. (C) Hematoxylin and eosin (H&E) staining of sections from the lung (top) and spleen (bottom) of Trp53-/- and Trp53Y217C/Y217C male mice, showing metastases in Trp53Y217C/Y217C animals. Normal organ structures are shown, with “A” indicating pulmonary alveoli, and “WP” and “RP” standing for splenic white and red pulp, respectively. In the lung section of the Trp53Y217C/Y217C mouse, the rectangle indicates a lymphoma area. In the spleen section of the Trp53Y217C/Y217C mouse, the typical splenic structures are absent due to massive tissue homogenization of the spleen by lymphoma cells.

We next performed a comparative analysis focusing on Trp53Y217C/Y217C and Trp53-/- thymi, which revealed 192 differentially expressed genes (Figure 5A-B and Table S4). An analysis of these data with GOrilla, the Gene Ontology enRIchment anaLysis and visuaLizAtion tool (Eden et al., 2009), indicated that 141 of these genes were associated with at least one gene ontology (GO) term, and revealed a significant enrichment for genes associated with white blood cell chemotaxis/migration and inflammation (Figures 5C and S11). Among these genes were notably Ccl17, Ccl9, Ccr3, Cxcl10, S100a8 and S100a9, six genes associated each with 10-15 GO terms related to white blood cell behavior and inflammation (Table S5). We next used RT-qPCR to quantify the expression of these genes in thymi from 8 weeks-old Trp53+/+, Trp53Y217C/Y217C and Trp53-/- male mice. Their expression was significantly increased in Trp53Y217C/Y217C thymic cells compared to Trp53-/-, or to both Trp53+/+ and Trp53-/- cells, consistent with gain or separation of function (GOF/SOF) effects in the Trp53Y217C mutant (Figure 5D). We also compared the transcriptomes of Trp53Y217C/Y217C and Trp53-/- thymocytes by gene set enrichment analysis (Subramanian et al., 2005). We found 13 gene sets significantly enriched in Trp53Y217C/Y217C cells with normalized enrichment scores (NES) > 2, among which three (“antimicrobial peptides”, “chemokine receptors bind chemokines” and “defensins”) were related to immunity (Figures 5E and S12A), consistent with an inflammatory response associated with the p53Y217C mutant. Other enriched gene sets notably included five sets (“electron transport chain”, “respiratory electron transport ATP synthesis by chemiosmotic coupling and heat production by uncoupling proteins”, “mitochondrial translation”, “respiratory electron transport” and “oxidative phosphorylation”) related to mitochondria function, and three sets (“deposition of new CENPA-containing nucleosomes at the centromere”, “arginine methyltransferases methylate histone arginines” and “PRC2 methylates histones and DNA”) related to chromatin plasticity (Figures 5E and S12B-C). These gene sets, differentially expressed between Trp53Y217C/Y217C and Trp53-/- cells, might contribute to accelerated tumorigenesis in Trp53Y217C/Y217C mice, given the reported impact of inflammation, metabolism changes and epigenetic plasticity on cancer evolution (Feinberg et al., 2006; Greten and Grivennikov, 2019; Kroemer and Pouyssegur, 2008). Furthermore, the hotspot mutants p53G245D or p53R273H were recently proposed to promote tumor progression by inducing non-canonical nuclear factor kappa light chain enhancer of activated B cell (NC-NF-κB) signaling (Zhao et al., 2024). In Trp53Y217C/Y217C cells, one gene set related to NC-NF-κB (“Dectin 1-mediated non-canonical NF-κB signaling”) was significantly enriched and two other NF-κB-related gene sets were also potentially enriched (Figure 5F and S13).

Evidence of inflammation in Trp53Y217C/Y217C mice.

(A-B) RNAseq analysis of thymi from Trp53Y217C/Y217C (n=3) and Trp53-/- (n=3) 8 weeks-old male mice. Volcano plot (A), with differentially expressed genes (DEGs), downregulated (blue) or upregulated (red) in Trp53Y217C/Y217C cells. Unsupervised clustering heat-map plot (B), with 192 DEGs ranked according to log2 fold changes. (C) Gene Ontology (GO) analysis of DEGs. Out of 192 DEGs, 141 are associated with at least one GO term, according to the Gene Ontology enRIchment anaLysis and visuaLizAtion tool (GOrilla). For each GO term, enrichment was calculated by comparing with the full list of 19759 genes associated with a GO term, and results are presented here with a color scale according to P-value of enrichment. This analysis mainly revealed an enrichment for genes associated with white blood cell chemotaxis/migration and inflammation, as shown here. Complete results of the analysis are presented in Figure S11. (D) RT-qPCR analysis of the indicated genes in thymi from 8 weeks-old Trp53+/+, Trp53Y217C/Y217C and Trp53-/- male mice. Means + s.e.m. (n=3 per genotype). ***P<0.001, **P<0.01, *P<0.05, °P<0.075 by Student’s t test. (E) Gene set enrichment analysis (GSEA) of transcriptomes from Trp53Y217C/Y217C and Trp53-/- thymic cells. GSEA identified 13 gene sets enriched in Trp53Y217C/Y217C cells with normalized enrichment scores (NES) > 2. Nominal p-values (NOM p-val) and false discovery rate q-values (FDR q-val) are indicated. (F) GSEA provides evidence of increased NC-NF-kB signaling in Trp53Y217C/Y217C male thymocytes. Results are presented as in E. (G) Supformin (LCC-12), an anti-inflammatory molecule, increases the female to male ratio in Trp53Y217C/Y217C weaned pups. Trp53+/Y217C females were mated with Trp53Y217C/Y217C males, then oral gavage of pregnant females with 5 mg/kg supformin was performed on the 10th and 11th days post-coitum, and their progeny was genotyped at weaning. The female to male (f/m) ratio for Trp53Y217C/Y217C weaned pups that were exposed to supformin in utero (+) was compared to the f/m ratios for Trp53-/- or Trp53Y217C/Y217C weaned pups never exposed to supformin (-). Exposure to supformin led to a five-fold increase in the proportion of Trp53Y217C/Y217C weaned females. **P<0.01, °P=0.056, ns: non-significant by Fischer’s test.

Reducing inflammation may rescue a fraction of female Trp53Y217C/Y217C embryos

Our data indicated that inflammation correlated with accelerated tumorigenesis in Trp53Y217C/Y217C male mice. Interestingly, inflammation was previously proposed to promote neural tube defects or embryonic death in a few mouse models (Lian et al., 2004; McNairn et al., 2019; Zhao et al., 2013). Furthermore CD44, encoding a cell-surface glycoprotein that drives inflammation and cancer progression (Solier et al., 2023), was recently identified as a key gene for the diagnosis and early detection of open neural tube defects (Karthik et al., 2022). Together, these data led us to hypothesize that, at least in a fraction of Trp53Y217C/Y217C females, neural tube defects might result from inflammation. To test this hypothesis, we mated Trp53+/Y217C female mice with Trp53Y217C/Y217C males, then administered supformin to pregnant females by oral gavage. Supformin (LCC-12), a potent inhibitor of the CD44-copper signaling pathway, was shown to reduce inflammation in vivo (Solier et al., 2023). Without any treatment, our previous crosses (Figure 3 and Table S1) collectively yielded a female to male ratio of 0.045 for Trp53Y217C/Y217C weaned mice (f/m=5/111 [4+1+0+0]/[75+22+9+5]), significantly different from the ratio of 0.29 (f/m=8/28) for Trp53-/- weaned animals. By contrast, in the progeny of supformin-treated pregnant mice we observed a five-fold increase in the female to male ratio for Trp53Y217C/Y217C weaned animals, to reach a value of 0.23 (f/m=3/13) indistinguishable from the ratio in Trp53-/- weaned animals from untreated mice (Figure 5G and Table S6). This result suggests that reducing inflammation in developing Trp53Y217C/Y217C female embryos may increase their viability.

Discussion

In this report, we generated a mouse model with a Trp53Y217C allele, the murine equivalent of the human hotspot mutant TP53Y220C. The analysis of this mouse model indicated that the p53Y217C mutation leads to the loss of many canonical wildtype p53 activities, notably the capacity to transactivate target genes important for tumor suppression (e.g. Bbc3, Cdkn1a and Zmat3). The fact that p53Y217C can rescue the embryonic lethality caused by a deficiency in Mdm2 or Mdm4 is also consistent with a major loss of canonical wildtype p53 functions (LOF) in this mutant. These findings are consistent with the notion that hotspot p53 mutants cause a complete or near complete LOF (Funk et al., 2025).

In addition, our analyses provide evidence that the p53Y217C mutation exhibits oncogenic effects. The possibility that a mutant p53 might acquire oncogenic functions was first suggested over 30 years ago (Dittmer et al., 1993), but this notion became controversial in recent years. Indeed, studies of a few p53 mutants in human acute myeloid leukemia and a mouse model of B cell lymphoma indicated that their increased tumorigenicity might result from dominant-negative effects (DNE) rather than gain of function (GOF), and a recent study further suggested that the putative GOF of many p53 mutants would not be required to sustain the expansion of tumors (Aubrey et al., 2018; Boettcher et al., 2019; Wang et al., 2023). Here, however, we observed accelerated tumorigenesis in Trp53Y217C/Y217C mice, but did not observe any evidence of DNE for spontaneous tumorigenesis in Trp53+/Y217C animals. Importantly, the fact that p53Y217C did not exhibit a DNE over the WT protein is consistent with the report that Li-Fraumeni patients carrying an heterozygous TP53Y220C mutation display a similar age of cancer onset than Li-Fraumeni patients with an heterozygous non-sense TP53 mutation (Xu et al., 2014). Furthermore, our evidence of an oncogenic GOF leading to aggressive metastatic tumors in Trp53Y217C/Y217C male mice is consistent with the oncogenic GOF attributed to p53Y220C in male patients with glioblastoma (Rockwell et al., 2021) or with experiments suggesting that p53Y220C expression in p53-null cells, or its overexpression in MCF-10A cells, may increase their migratory or invasive capacities (Pal et al., 2023; Zhou et al., 2022).

Our transcriptomic analyses suggested that the p53Y217C mutant might accelerate tumorigenesis by promoting inflammation, a hallmark of cancer (Hanahan, 2022) previously associated with a few other hotspot p53 mutants with oncogenic GOF (Agupitan et al., 2020; Behring et al., 2019; Ham et al., 2019; Zhao et al., 2024). Interestingly, the GSEA comparison of transcriptomes from Trp53Y217C/Y217C and Trp53-/- male thymic cells indicated differences related to inflammation and chemokine signaling, but also to mitochondria function and chromatin plasticity, and a link between inflammation, mitochondrial copper and epigenetic plasticity was recently demonstrated (Solier et al., 2023). Furthermore, the p53G245D or p53R273H mutants were proposed to accelerate tumorigenesis by promoting a chromosomal instability that would activate NC-NF-kB signaling and promote tumor cell metastasis (Zhao et al., 2024). Here we observed increased chromosomal instability in Trp53Y217C/Y217C fibroblasts, increased NC-NF-kB and inflammation-related signaling in Trp53Y217C/Y217C thymocytes, and increased tumor metastasis in Trp53Y217C/Y217C mice compared to their Trp53-/- counterparts. These data suggest that similar mechanisms might underlie the oncogenic properties of the p53Y217C, p53G245D and p53R273H mutants.

A striking and unexpected effect of the p53Y217C mutation was its severe impact on the perinatal viability of female mice, which led to observe only four Trp53Y217C/Y217C females for 75 Trp53Y217C/Y217C males at weaning from heterozygous intercrosses, or five Trp53Y217C/Y217C females for 111 Trp53Y217C/Y217C males when all weaned animals from relevant crosses were considered. The perinatal lethality of Trp53Y217C/Y217C females correlated with a high frequency (38.5%) of exencephalic females at E12.5-E16.5 embryonic ages. By comparison, the female to male ratio at weaning was 8/28 for Trp53-/- mice, and only 0-8% of exencephalic female Trp53-/- embryos were reported in similar genetic backgrounds (Armstrong et al., 1995; Donehower et al., 1992; Sah et al., 1995). These observations provide evidence of teratogenic effects for the p53Y217C mutant protein. Three other mouse p53 models were previously found to cause exencephaly at a higher frequency than a p53 null allele: Trp53NLS1, with 3 mutations at residues 316-318 affecting a nuclear localization signal and leading to a predominantly cytoplasmic localization of p53NLS1 in most cells (Regeling et al., 2011); Trp53N236S, a mouse model of TP53N239S, a recurrent but uncommon mutant in human cancers (Zhao et al., 2019); and Bim+/- Trp53-/- mice, combining p53 loss with an haploinsufficiency in the cytoplasmic proapoptotic regulator Bim/Bcl2l11 (Delbridge et al., 2019). The female-specific lethality of Trp53N236S mice was proposed to result from increased Xist expression, whereas a decrease in Xist expression was observed in Bim+/- Trp53-/- mice, suggesting distinct causes for neural tube defects (Delbridge et al., 2019; Zhao et al., 2019). We observed decreased Xist expression and an increased expression of X-linked genes in neurospheres from Trp53Y217C/Y217C females compared to WT females, whereas these genes were expressed at similar levels in neurospheres from Trp53Y217C/Y217C and Trp53-/- females. These results are not consistent with the mechanism proposed for exencephaly in the Trp53N236S mouse model, but might rather reflect a p53 LOF in Trp53Y217C/Y217C embryos. In addition, the analyses of the Trp53NLS1 and Bim+/- Trp53-/- mouse models make it tempting to speculate that, in female Trp53Y217C/Y217C embryos, the abundance of p53Y217C in the cytoplasm might perturb mitochondria function (Blandino et al., 2020) to further promote neural tube closure defects. Consistent with this possibility, p53 mutant proteins accumulating in the cytoplasm were reported to inhibit autophagy (Morselli et al., 2008), a key determinant for mitochondrial integrity (Rambold and Lippincott-Schwartz, 2011), and autophagy impairment may promote neural tube defects (Fimia et al., 2007; Ye et al., 2020).

Importantly, our data suggest that common mechanisms might contribute to the oncogenic effects accelerating tumorigenesis in Trp53Y217C/Y217C males and the teratogenic effects causing the perinatal lethality of many Trp53Y217C/Y217C females. Indeed, the RNAseq analysis of male thymocytes indicated that inflammation likely promotes oncogenesis in Trp53Y217C/Y217C males, and the administration of the anti-inflammatory drug supformin to pregnant mice increased the ratio of Trp53Y217C/Y217C weaned females in their progeny. Thus, the low female/male ratio observed for Trp53Y217C/Y217C weaned animals would likely result not only from aberrant X-chromosome inactivation in female embryos as in Trp53-/- animals (Delbridge et al., 2019), but also from inflammation. Presumably, a higher level of chromosomal instability in Trp53Y217C/Y217C cells might promote inflammation in Trp53Y217C/Y217C embryos (Rodier et al., 2009), and this might be more detrimental to females because Trp53Y217C/Y217C males would be protected by the anti-inflammatory effects of androgens (Ainslie et al., 2024; Bianchi, 2019; McNairn et al., 2019). Of note, we also observed that decreased Mdm4 levels increased the female/male ratio in Trp53Y217C/Y217C weaned mice. Because MDM4 overexpression was shown to promote genome instability and inflammation in cells independently of p53 (Carrillo et al., 2015; Liu et al., 2024), it is tempting to speculate that, as for supformin-treated mice, the improved survival of Trp53Y217C/Y217C female embryos in an Mdm4+/- background might result from decreased inflammation. Interestingly, in the offspring of human pregnancies at risk of early preterm delivery, an Interleukin-6 polymorphism was associated with a female-specific susceptibility to adverse neurodevelopmental outcome (Varner et al., 2020), suggesting that inflammation may also impact the development of human female embryos. Finally, our mating attempts with Trp53Y217C/Y217C mice indicated that Trp53Y217C/Y217C pregnant females exhibited dystocia, and parturition is considered to be a finely regulated inflammatory process (Zhang and Wei, 2021). Of note, dystocia was not observed in Trp53-/- mice, but was previously reported in pregnant females with a combined deficiency of p53 and FasL (Embree-Ku and Boekelheide, 2002). The fact that Trp53Y217C/Y217C females shared phenotypic traits with both Bim+/- Trp53-/- mice (neural tube defects) and FasL-/- Trp53-/- mice (dystocia) may seem relevant, because Bim and FasL are both regulators of apoptosis and autophagy that also regulate immune responses (Sionov et al., 2015; Taylor and Ng, 2018). Altogether, these data suggest that inflammation may cause accelerated tumorigenesis in Trp53Y217C/Y217C male mice and contribute to promote neural tube defects in Trp53Y217C/Y217C female embryos or dystocia in Trp53Y217C/Y217C pregnant females (Table S7).

In conclusion, we generated a mouse model expressing p53Y217C, the murine equivalent of human p53Y220C, and its analysis revealed that the gain-of-function of a hotspot mutant p53 may not be limited to oncogenic effects. This is an important notion to consider given the ever-expanding functions ascribed to p53 (Hu, 2009; Rakotopare and Toledo, 2023; Voskarides and Giannopoulou, 2023; Vousden and Lane, 2007). In addition, our results unveiled the potential relevance of supformin in obstetrics, which deserves further investigation.

Materials and methods

LSL-Y217C construct

We used mouse genomic Trp53 DNA from previous constructs, containing a portion of intron 1 with a LoxP-Stop-LoxP (LSL) cassette (Ventura et al., 2007) in which the puromycin-resistance gene was replaced by a neomycin-resistance gene (Simeonova et al., 2013). The point mutation (A to G) encoding a Tyr to Cys substitution (TAT to TGT) at codon 217, together with a silent mutation nearby creating a Ban II restriction site, were introduced by PCR directed mutagenesis by using primers Y217C-F and Y217C-R (See Table S8 for primer sequences). The resulting Trp53LSL-Y217C targeting vector was fully sequenced before use.

Targeting in ES Cells and Genotyping

129/SvJ embryonic stem (ES) cells were electroporated with the targeting construct linearized with Not I. Two recombinant clones, identified by long-range PCR and confirmed by internal PCR and Southern Blot, were injected into C57BL/6J blastocysts to generate chimeras, and germline transmission was verified by genotyping their offspring. In vivo excision of the LSL cassette was performed by breeding Trp53+/LSL-Y217C male mice with females carrying the PGK-Cre transgene (Lallemand et al., 1998) and genotyping their offspring. The Trp53Y217C (p53YC) mutation was routinely genotyped by PCR using primers c and d (Figure 1, Table S8) followed by BanII digestion. Tumorigenesis studies were performed on mouse cohorts resulting from ≥ 5 generations of backcrosses with C57BL6/J mice. For all experiments, mice housing and treatment were conducted according to the Institutional Animal Care and Use Committee of the Institut Curie (approved project #03769.02).

Cells and Cell Culture Reagents

Mouse embryonic fibroblasts (MEFs) isolated from 13.5-day embryos were cultured in a 5% CO2 and 3% O2 incubator, in DMEM GlutaMAX (Gibco), with 15% FBS (Biowest), 100 μM 2-mercaptoethanol (Millipore), 0.01 mM Non-Essential Amino-Acids, and penicillin/streptomycin (Gibco) for ≤ 6 passages. Cells were irradiated with a Cs γ-irradiator or treated for 24 hr with 10 μM Nutlin 3a (Sigma-Aldrich) or 1 μM Doxorubicin (Sigma-Aldrich). Neurospheres were prepared by isolating neural stem/progenitor cells (NPCs) from the lateral and median ganglionic eminences of E14.5 embryos. NPCs were cultured in a 5% CO2 incubator, in DMEM/F12 medium (with L-Glutamine, Invitrogen) supplemented with 1% B-27 (Invitrogen), 1% N2 (Invitrogen), 0.3% glucose (Invitrogen), 25 μg/ml insulin (Sigma-Aldrich), 20 ng/ml EGF (PeproTech), 10 ng/ml bFGF (PeproTech) and penicillin/streptomycin (Gibco). Neurospheres were dissociated once a week by Accutase treatment and mechanically, then seeded at 106 cells per 6 cm diameter dish.

Quantitative RT-PCR

Total RNA, extracted using NucleoSpin RNA II (Macherey-Nagel), was reverse transcribed using SuperScript IV (Invitrogen). Real-time quantitative PCRs were performed as previously described (Simeonova et al., 2012), on an ABI PRISM 7500 using Power SYBR Green (Applied Biosystems). Primer sequences are reported in Table S8.

Western Blots

Cells were lysed in RIPA buffer (50 mM Tris-HCl pH 8, 150 mM NaCl, 5 mM EDTA, 0.5% deoxycholic acid, 0.1% SDS, 1% NP-40) supplemented with Protease inhibitors cocktail (Roche) and 1 mM PMSF (Sigma). Whole-cell extracts were sonicated three times for 10 s and centrifuged at 13 000 rpm for 30 min to remove cell debris. Protein lysate concentration was determined by BCA assay (Thermo Scientific) and 30 μg of each lysate was fractionated by SDS–PAGE on a 4-12% polyacrylamide gel and transferred onto polyvynilidene difluoride (PVDF) membranes (Amersham). Membranes were incubated with antibodies raised against Mdm2 (MDM2-4B2, Abcam, 1/500), p53 (CM-5, Novocastra, 1/2000), p21 (F5, Santa Cruz Biotechnology, 1/200) and actin (actin-HRP sc47778, Santa Cruz Biotechnology, 1/5000). Chemiluminescence revelation of western blots was achieved with the SuperSignal West Dura (Perbio).

ChIP Assay

ChIP analysis was performed as previously described (Simeonova et al., 2013). p53-DNA complexes were immunoprecipitated from total extracts by using 50 μg of a polyclonal antibody against p53 (FL-393, Santa Cruz Biotechnology) and 300 μg of sonicated chromatin. Rabbit IgG (Abcam) was used for control precipitation. Quantitative PCR was performed on ABI PRISM 7500. Primer sequences are reported in Table S8.

Cell Fractionation Assay

Cells were lysed in hypotonic lysis buffer (10 mM Tris pH 7.5, 10 mM NaCl, 3 mM Mgcl2, 0.3% NP-40, 10% glycerol) on ice for 10 min, then centrifuged at 4°C for 3 min at 1000g. The supernatant was recovered as the cytoplasmic fraction, and the pellet was incubated in a modified Wuarin-Schibler buffer (10 mM Tris-HCl pH 7.0, 4 mM EDTA, 0.3 M NaCl, 1 M urea, 1% NP-40) on ice for 5 min, vortexed, incubated on ice for 10 min, then centrifuged at 4°C for 3 min at 1000g. The supernatant was recovered as the nucleoplasmic fraction, and the pellet was washed, then incubated in a nuclear lysis buffer (20 mM Tris pH 7.5, 150 mM KCl, 3 mM MgCl2, 0.3% NP-40, 10% glycerol) on ice for 5 min, sonicated, then centrifuged at 13000 rpm at 4°C and the supernatant was recovered as the chromatin fraction. All buffers were supplemented with Protease inhibitors cocktail (Roche). Cellular fractions were analyzed by western blots with antibodies against p53 (AF-1355, R&D systems, 1/600), Tubulin (ab15568, Abcam, 1/1000), Nup98 (ab50610, Abcam, 1/1000) and histone H3 (ab1791, Abcam, 1/1000). Chemiluminescence revelation was achieved with SuperSignal West Pico or Femto (Thermoscientific) and analyzed with a ChemiDoc imaging system (Bio-Rad).

Immunofluorescence

MEFs were cultured on collagen-coated coverslips, treated with 10 μM Nutlin 3a (Sigma-Aldrich) for 24 hr and analyzed. Coverslips were stained with rabbit anti-p53 FL-393 (Santa Cruz Biotechnology, 1/50), mouse anti-actin primary antibody and with Alexa Fluor 647 anti-Rabbit and Alexa Fluor 488 anti-Mouse secondary antibodies (Molecular Probes). DNA was counterstained with DAPI. Images were captured on an epifluorescence microscope using equal exposure times for all images for each fluor.

Cell-Cycle Assay

Log phase cells were irradiated at room temperature (RT) with a Cs γ-irradiator at doses of 3 or 12 Gy, incubated for 24 hr, then pulse labeled for 1 hr with BrdU (10 μM), fixed in 70% ethanol, double stained with FITC anti-BrdU and propidium iodide, and analyzed by flow cytometry with a BD Biosciences FACSort and the FlowJo software.

Apoptosis Assay

Six to eight weeks-old mice were left untreated or submitted to 10 Gy whole-body ɣ-irradiation. Mice were sacrificed 4 hr later and thymi were extracted. Thymocytes were recovered by filtration through a 70 μm smash, stained with Annexin V-FITC Apoptosis detection kit (Abcam) and propidium iodide, then analyzed with a LSRII FACS machine. Data were analyzed using Flowjo software.

Metaphase spread preparation and analysis

Primary MEFs (at passage 4 for all genotypes) were treated with 0.1 mM nocodazole for 3 h to arrest cells in metaphase, then submitted to hypotonic shock (75 mM KCl), fixed in a (3:1) ethanol/acetic acid solution, dropped onto glass slides, then air-dried slides were stained with Giemsa to score for chromosome aberrations. Images were acquired using a Zeiss Axiophot (X63) microscope.

Histology

Organs were fixed in formol 4% for 24 hr, then ethanol 70%, and embedded in paraffin wax. Serial sections of 3 μm were stained as described (Simeonova et al., 2013), with hematoxylin and eosin using standard procedures.

RNA-Seq analysis

Total RNA was extracted from the thymi of 8 weeks-old asymptomatic male mice using nucleospin RNA II (Macherey-Nagel). The quality of RNA was checked with Bioanalyzer Agilent 2100 and RNAs with a RIN (RNA integrity number) ≥ 7 were retained for further analysis. RNA was depleted from ribosomal RNA, then converted into cDNA libraries using a TruSeq Stranded Total Library preparation kit (Illumina). Paired-end sequencing was performed on an Illumina MiSeq platform. Reads were mapped to the mouse genome version GRCm38 and counted on gene annotation gencode.vM18 with featureCounts (Liao et al., 2014). Differentially expressed genes with an adjusted p-value < 0.05 were identified using the DESeq2 R package (Love et al., 2014).

Gene Ontology analysis

Gene ontology analysis of differentially expressed genes was performed by using the GOrilla (Technion) software as previously described (Rakotopare et al., 2023). Enrichment analyses were carried out by comparing the list of 192 differentially expressed genes between Trp53Y217C/Y217C and Trp53-/- thymi to the full list of genes (background), with ontology searches for biological processes and default P-value settings (10−3).

Gene set enrichment analysis

Gene set enrichment analysis was performed by using the GSEA software with canonical pathway gene sets from the Mouse Molecular Signature Database (MSigDB) (Subramanian et al., 2005). Gene sets with a false discovery rate (FDR) q-value < 0.25 were regarded as potentially relevant. The best candidates, with an FDR q-value < 0.01, exhibited normalized enrichment scores (NES) > 1.91 and nominal (NOM) P-values <0.003.

Oral gavage of pregnant females

Trp53+/Y217C female mice were mated with Trp53Y217C/Y217C males, then mice with a vaginal plug on the next day were separated from males and weighted every other day to confirm pregnancy. On the 10th and 11th days post-coitum, pregnant mice received 5 mg/kg of supformin (in a PBS solution) by oral gavage with flexible gauge feeding tubes, then their pups were genotyped at weaning.

Statistical analyses

Unless noted otherwise, we used a Student’s t test to analyze differences between two groups of values. For the proportions of mice at weaning, the observed mouse count was compared to the expected count according to Mendel’s distribution and a Chi-square (χ2) test, and a Fischer’s test was applied to compare frequencies of mutant females or female to male ratios. Fischer’s tests were also used to analyze frequencies of chromosome rearrangements in fibroblasts. Survival curves were analyzed with log-rank Mantel-Cox tests. Analyses were performed by using GraphPad Prism, and values of P ≤ 0.05 were considered significant.